### 2024-09-19 paper/paper.qmd
added sketches, cleaned data, added graphs into paper

---
title: "Week 3 Reflection"
subtitle: "Marriage Data Cleaning"
author: "Cher Ning-Li"
thanks: "Code and data are available at: LINK."
date: today
date-format: long
abstract: "In this paper I analyze the number of marriages across the years, from 2018-2023."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)

data <- read_csv(here::here("data/analysis_data/analysis_data.csv"))
```

# Introduction

You can and should cross-reference sections and sub-sections. We use @citeR and @rohan.

The remainder of this paper is structured as follows. @sec-data....

# Data {#sec-data}

Some of our data is of penguins (@fig-marriages), from @OpenDataToronto.

```{r}
#| label: fig-marriages
#| fig-cap: Number of Marriages From 2018-2023
#| echo: false

data |> 
  ggplot(aes(x = date, y = marriage_licenses)) + 
  geom_point()
```

We can see that the number of marriages fluctuates across the years.

# Model

The goal of our modelling strategy is twofold. Firstly,...

Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].

## Model set-up

Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.

\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_i + \gamma_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}

We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.

### Model justification

We expect a positive relationship between the size of the wings and time spent aloft. In particular...

We can use maths by including latex between dollar signs, for instance $\theta$.

# Results

Our results are summarized in @tbl-modelresults.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

library(rstanarm)

first_model <-
  readRDS(file = here::here("models/first_model.rds"))
```

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
#| warning: false

modelsummary::modelsummary(
  list(
    "First model" = first_model
  ),
  statistic = "mad",
  fmt = 2
)
```

# Discussion

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

## Second discussion point

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\newpage

\appendix

# Appendix {-}

# Additional data details

# Model details {#sec-model-details}

## Posterior predictive check

In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...

In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckandposteriorvsprior
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits, and is affected by, the data"
#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]

pp_check(first_model) +
  theme_classic() +
  theme(legend.position = "bottom")

posterior_vs_prior(first_model) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom") +
  coord_flip()
```

## Diagnostics

@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...

@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot", "Rhat plot"]
#| layout-ncol: 2

plot(first_model, "trace")

plot(first_model, "rhat")
```

\newpage

# References

### 2024-09-19 scripts/00-simulate_data.R

@@ -1,19 +1,40 @@
 #### Preamble ####
-# Purpose: Simulates... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose: Simulates data
+# Author: Cher Ning-Li
+# Date: 19 September 2024
+# Contact: cher.ning@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Pre-requisites: None
+# Any other information needed? None
 
 
 #### Workspace setup ####
 library(tidyverse)
+
 # [...UPDATE THIS...]
 
 #### Simulate data ####
-# [...ADD CODE HERE...]
+set.seed(304)
 
+# Define the start and end date
+start_date <- as.Date("2018-01-01")
+end_date <- as.Date("2023-12-31")
 
+# Set the number of random dates you want to generate
+number_of_dates <- 100
 
+data <-
+  tibble(
+    dates = as.Date(
+      runif(
+        n = number_of_dates,
+        min = as.numeric(start_date),
+        max = as.numeric(end_date)
+      ),
+      origin = "1970-01-01"
+    ),
+    number_of_marriage = rpois(n = number_of_dates, lambda = 10)
+  )
+
+# Write csv
+write_csv(data, file = "data/raw_data/simulated.csv")

### 2024-09-19 scripts/01-download_data.R

@@ -1,8 +1,8 @@
 #### Preamble ####
 # Purpose: Downloads and saves the data from [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Author: Cher Ning-Li
+# Date: 19 September 2024
+# Contact: cher.ning@mail.utoronto.ca
 # License: MIT
 # Pre-requisites: [...UPDATE THIS...]
 # Any other information needed? [...UPDATE THIS...]
@@ -11,16 +11,24 @@
 #### Workspace setup ####
 library(opendatatoronto)
 library(tidyverse)
-# [...UPDATE THIS...]
 
 #### Download data ####
-# [...ADD CODE HERE TO DOWNLOAD...]
 
+# get package
+package <- show_package("e28bc818-43d5-43f7-b5d9-bdfb4eda5feb")
+
+# get all resources for this package
+resources <- list_package_resources("e28bc818-43d5-43f7-b5d9-bdfb4eda5feb")
+
+# identify datastore resources; by default, Toronto Open Data sets datastore resource format to CSV for non-geospatial and GeoJSON for geospatial resources
+datastore_resources <- filter(resources, tolower(format) %in% c('csv', 'geojson'))
+
+# load the first datastore resource as a sample
+data <- filter(datastore_resources, row_number()==1) %>% get_resource()
 
 
 #### Save data ####
-# [...UPDATE THIS...]
 # change the_raw_data to whatever name you assigned when you downloaded it.
-write_csv(the_raw_data, "inputs/data/raw_data.csv") 
+write_csv(data, "data/raw_data/raw_data.csv") 
 
          

### 2024-09-19 scripts/02-data_cleaning.R

@@ -1,44 +1,27 @@
 #### Preamble ####
 # Purpose: Cleans the raw plane data recorded by two observers..... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 6 April 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Author: Cher Ning-Li
+# Date: 19 September 2024
+# Contact: cher.ning@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Pre-requisites: Need to have downloaded data
+# Any other information needed? None
 
 #### Workspace setup ####
 library(tidyverse)
 
 #### Clean data ####
-raw_data <- read_csv("inputs/data/plane_data.csv")
+raw_data <- read_csv("data/raw_data/raw_data.csv")
 
 cleaned_data <-
   raw_data |>
   janitor::clean_names() |>
-  select(wing_width_mm, wing_length_mm, flying_time_sec_first_timer) |>
-  filter(wing_width_mm != "caw") |>
-  mutate(
-    flying_time_sec_first_timer = if_else(flying_time_sec_first_timer == "1,35",
-                                   "1.35",
-                                   flying_time_sec_first_timer)
-  ) |>
-  mutate(wing_width_mm = if_else(wing_width_mm == "490",
-                                 "49",
-                                 wing_width_mm)) |>
-  mutate(wing_width_mm = if_else(wing_width_mm == "6",
-                                 "60",
-                                 wing_width_mm)) |>
-  mutate(
-    wing_width_mm = as.numeric(wing_width_mm),
-    wing_length_mm = as.numeric(wing_length_mm),
-    flying_time_sec_first_timer = as.numeric(flying_time_sec_first_timer)
-  ) |>
-  rename(flying_time = flying_time_sec_first_timer,
-         width = wing_width_mm,
-         length = wing_length_mm
-         ) |> 
-  tidyr::drop_na()
+  
+  separate(col = time_period, 
+            into = c("year", "month"),
+            sep = "-") |>
+  mutate(date = lubridate::ymd(paste(year, month, "01", sep = "-")))
+
 
 #### Save data ####
-write_csv(cleaned_data, "outputs/data/analysis_data.csv")
+write_csv(cleaned_data, "data/analysis_data/analysis_data.csv")

### 2024-09-19 scripts/03-test_data.R

@@ -1,15 +1,21 @@
 #### Preamble ####
-# Purpose: Tests... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose: Tests data
+# Author: Cher Ning-Li
+# Date: 19 September 2024
+# Contact: cher.ning@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Pre-requisites: Must have data downloaded and cleaned
+# Any other information needed? None
 
 
 #### Workspace setup ####
 library(tidyverse)
-# [...UPDATE THIS...]
 
 #### Test data ####
+data <- read_csv("data/raw_data/simulated.csv")
+
+# Test there are no negative number of marriages
+data$number_of_marriage |> min() <= 0
+
+# Test for NAs
+all(is.na(data$number_of_marriage))

### 2024-09-19 paper/paper.qmd
removed unrelated content

@@
 
-Some of our data is of penguins (@fig-marriages), from @OpenDataToronto.
+Some of our data is of marriages (@fig-marriages), from @OpenDataToronto.
 
@@
 
-The goal of our modelling strategy is twofold. Firstly,...
-
-Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].
-
 ## Model set-up
-
-Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.
-
-\begin{align} 
-y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
-\mu_i &= \alpha + \beta_i + \gamma_i\\
-\alpha &\sim \mbox{Normal}(0, 2.5) \\
-\beta &\sim \mbox{Normal}(0, 2.5) \\
-\gamma &\sim \mbox{Normal}(0, 2.5) \\
-\sigma &\sim \mbox{Exponential}(1)
-\end{align}
 
@@
 ### Model justification
-
-We expect a positive relationship between the size of the wings and time spent aloft. In particular...
-
-We can use maths by including latex between dollar signs, for instance $\theta$.
 
@@
 
-first_model <-
-  readRDS(file = here::here("models/first_model.rds"))
 ```
@@
 
-If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.
-
 ## Second discussion point
@@
 ## Weaknesses and next steps
-
-Weaknesses and next steps should also be included.
 
@@
 
-In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...
-
-In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...
-
-```{r}
-#| eval: true
-#| echo: false
-#| message: false
-#| warning: false
-#| label: fig-ppcheckandposteriorvsprior
-#| layout-ncol: 2
-#| fig-cap: "Examining how the model fits, and is affected by, the data"
-#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]
-
-pp_check(first_model) +
-  theme_classic() +
-  theme(legend.position = "bottom")
-
-posterior_vs_prior(first_model) +
-  theme_minimal() +
-  scale_color_brewer(palette = "Set1") +
-  theme(legend.position = "bottom") +
-  coord_flip()
-```
-
 ## Diagnostics
-
-@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...
-
-@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...
-
-```{r}
-#| echo: false
-#| eval: true
-#| message: false
-#| warning: false
-#| label: fig-stanareyouokay
-#| fig-cap: "Checking the convergence of the MCMC algorithm"
-#| fig-subcap: ["Trace plot", "Rhat plot"]
-#| layout-ncol: 2
-
-plot(first_model, "trace")
-
-plot(first_model, "rhat")
-```
 

### 2024-09-24 scripts/01-download_data.R
download new data set

@@ -1,31 +1,29 @@
 #### Preamble ####
-# Purpose: Downloads and saves the data from [...UPDATE THIS...]
+# Purpose: Downloads and saves the data from Toronto Open Data
 # Author: Cher Ning-Li
-# Date: 19 September 2024
+# Date: 24 September 2024
 # Contact: cher.ning@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Pre-requisites: None
+# Any other information needed? None
 
 
 #### Workspace setup ####
 library(opendatatoronto)
-library(tidyverse)
-
-#### Download data ####
+library(dplyr)
 
 # get package
-package <- show_package("e28bc818-43d5-43f7-b5d9-bdfb4eda5feb")
+package <- show_package("a7ae08f3-c512-4a88-bb3c-ab40eca50c5e")
+package
 
 # get all resources for this package
-resources <- list_package_resources("e28bc818-43d5-43f7-b5d9-bdfb4eda5feb")
+resources <- list_package_resources("a7ae08f3-c512-4a88-bb3c-ab40eca50c5e")
 
 # identify datastore resources; by default, Toronto Open Data sets datastore resource format to CSV for non-geospatial and GeoJSON for geospatial resources
 datastore_resources <- filter(resources, tolower(format) %in% c('csv', 'geojson'))
 
 # load the first datastore resource as a sample
 data <- filter(datastore_resources, row_number()==1) %>% get_resource()
-
 
 #### Save data ####
 # change the_raw_data to whatever name you assigned when you downloaded it.

### 2024-09-24 scripts/01-download_data.R
switched to shelters dataset because previous was too small

@@ -11,13 +11,14 @@
 #### Workspace setup ####
 library(opendatatoronto)
 library(dplyr)
+# package to use write_csv
+library(readr)
 
 # get package
-package <- show_package("a7ae08f3-c512-4a88-bb3c-ab40eca50c5e")
-package
+package <- show_package("21c83b32-d5a8-4106-a54f-010dbe49f6f2")
 
 # get all resources for this package
-resources <- list_package_resources("a7ae08f3-c512-4a88-bb3c-ab40eca50c5e")
+resources <- list_package_resources("21c83b32-d5a8-4106-a54f-010dbe49f6f2")
 
 # identify datastore resources; by default, Toronto Open Data sets datastore resource format to CSV for non-geospatial and GeoJSON for geospatial resources
 datastore_resources <- filter(resources, tolower(format) %in% c('csv', 'geojson'))

### 2024-09-24 scripts/00-simulate_data.R
new simulated data

@@ -1,7 +1,7 @@
 #### Preamble ####
 # Purpose: Simulates data
 # Author: Cher Ning-Li
-# Date: 19 September 2024
+# Date: 24 September 2024
 # Contact: cher.ning@mail.utoronto.ca
 # License: MIT
 # Pre-requisites: None
@@ -10,8 +10,6 @@
 
 #### Workspace setup ####
 library(tidyverse)
-
-# [...UPDATE THIS...]
 
 #### Simulate data ####
 set.seed(304)
@@ -33,7 +31,7 @@ data <-
       ),
       origin = "1970-01-01"
     ),
-    number_of_marriage = rpois(n = number_of_dates, lambda = 10)
+    number_of_vacancies = rpois(n = number_of_dates, lambda = 10)
   )
 
 # Write csv

### 2024-09-24 paper/paper.qmd
updated some references

@@
 ---
-title: "Week 3 Reflection"
-subtitle: "Marriage Data Cleaning"
+title: "Daily Shelter and Overnight Services Availabilities"
+#subtitle: "Marriage Data Cleaning"
 author: "Cher Ning-Li"
-thanks: "Code and data are available at: LINK."
+thanks: "Code and data are available at: https://github.com/cher-ning/shelter_occupancies"
 date: today
 date-format: long
-abstract: "In this paper I analyze the number of marriages across the years, from 2018-2023."
+abstract: "In this paper I analyze the capacity vs real occupancy of shelters across Canada."
+#TODO Complete abstract after paper.
 format: pdf
@@
 
-You can and should cross-reference sections and sub-sections. We use @citeR and @rohan.
+You can and should cross-reference sections and sub-sections. We use @citeR and @citeTidyverse
 
@@
 
-Some of our data is of marriages (@fig-marriages), from @OpenDataToronto.
+Some of our data is of marriages (@fig-marriages), from @opendatatoronto.
 

### 2024-09-24 scripts/01-download_data.R
updated new dataset sketch

@@ -28,6 +28,4 @@ data <- filter(datastore_resources, row_number()==1) %>% get_resource()
 
 #### Save data ####
 # change the_raw_data to whatever name you assigned when you downloaded it.
-write_csv(data, "data/raw_data/raw_data.csv") 
-
-         
+write_csv(data, "data/raw_data/raw_data.csv")

### 2024-09-24 scripts/00-simulate_data.R
new simulated data based on new sketch

@@ -31,7 +31,8 @@ data <-
       ),
       origin = "1970-01-01"
     ),
-    number_of_vacancies = rpois(n = number_of_dates, lambda = 10)
+    capacity = rpois(n = number_of_dates, lambda = 100),
+    vacancies = rpois(n = number_of_dates, lambda = 10)
   )
 
 # Write csv

### 2024-09-24 scripts/02-data_cleaning.R
filtered out unneeded variables and rows with NA data, selected for rows of interest (toronto emergency shelters with bed based occupancy)

@@ -1,5 +1,5 @@
 #### Preamble ####
-# Purpose: Cleans the raw plane data recorded by two observers..... [...UPDATE THIS...]
+# Purpose: Cleans the raw shelter data for only relevant variables
 # Author: Cher Ning-Li
 # Date: 19 September 2024
 # Contact: cher.ning@mail.utoronto.ca
@@ -13,15 +13,19 @@ library(tidyverse)
 #### Clean data ####
 raw_data <- read_csv("data/raw_data/raw_data.csv")
 
-cleaned_data <-
-  raw_data |>
-  janitor::clean_names() |>
-  
-  separate(col = time_period, 
-            into = c("year", "month"),
-            sep = "-") |>
-  mutate(date = lubridate::ymd(paste(year, month, "01", sep = "-")))
+# filters for only Bed Based and Emergency type shelters located in Toronto
+filtered <- subset(data, CAPACITY_TYPE == 'Bed Based Capacity' &
+                     PROGRAM_MODEL == 'Emergency' & LOCATION_CITY == 'Toronto')
 
+# removes unnecessarily variables
+filtered <- filtered %>% select(OCCUPANCY_DATE, SHELTER_ID, 
+                                LOCATION_POSTAL_CODE, CAPACITY_ACTUAL_BED,
+                                CAPACITY_FUNDING_BED, OCCUPIED_BEDS,
+                                UNOCCUPIED_BEDS, UNAVAILABLE_BEDS, 
+                                OCCUPANCY_RATE_BEDS)
+
+# removes rows with NA data
+cleaned_data <- filtered %>% drop_na()
 
 #### Save data ####
 write_csv(cleaned_data, "data/analysis_data/analysis_data.csv")

### 2024-09-25 scripts/01-download_data.R
updated data download lines to bypass previous 32,000 capacity

@@ -11,21 +11,12 @@
 #### Workspace setup ####
 library(opendatatoronto)
 library(dplyr)
+library(tidyverse)
 # package to use write_csv
 library(readr)
 
-# get package
-package <- show_package("21c83b32-d5a8-4106-a54f-010dbe49f6f2")
-
-# get all resources for this package
-resources <- list_package_resources("21c83b32-d5a8-4106-a54f-010dbe49f6f2")
-
-# identify datastore resources; by default, Toronto Open Data sets datastore resource format to CSV for non-geospatial and GeoJSON for geospatial resources
-datastore_resources <- filter(resources, tolower(format) %in% c('csv', 'geojson'))
-
-# load the first datastore resource as a sample
-data <- filter(datastore_resources, row_number()==1) %>% get_resource()
+# accesses link directly to download full data file
+all_data <- read_csv("https://ckan0.cf.opendata.inter.prod-toronto.ca/dataset/21c83b32-d5a8-4106-a54f-010dbe49f6f2/resource/ffd20867-6e3c-4074-8427-d63810edf231/download/Daily%20shelter%20overnight%20occupancy.csv")
 
 #### Save data ####
-# change the_raw_data to whatever name you assigned when you downloaded it.
-write_csv(data, "data/raw_data/raw_data.csv")
+write_csv(all_data, "data/raw_data/raw_data.csv")

### 2024-09-25 scripts/02-data_cleaning.R

@@ -14,7 +14,7 @@ library(tidyverse)
 raw_data <- read_csv("data/raw_data/raw_data.csv")
 
 # filters for only Bed Based and Emergency type shelters located in Toronto
-filtered <- subset(data, CAPACITY_TYPE == 'Bed Based Capacity' &
+filtered <- subset(raw_data, CAPACITY_TYPE == 'Bed Based Capacity' &
                      PROGRAM_MODEL == 'Emergency' & LOCATION_CITY == 'Toronto')
 
 # removes unnecessarily variables
@@ -24,8 +24,12 @@ filtered <- filtered %>% select(OCCUPANCY_DATE, SHELTER_ID,
                                 UNOCCUPIED_BEDS, UNAVAILABLE_BEDS, 
                                 OCCUPANCY_RATE_BEDS)
 
-# removes rows with NA data
-cleaned_data <- filtered %>% drop_na()
+# adds column for month
+# filtered$month <- format(as.Date(filtered$OCCUPANCY_DATE), "%m")
+
+# removes rows with NA data, cleans variables names
+cleaned_data <- filtered %>% drop_na() |> janitor::clean_names() 
+
 
 #### Save data ####
 write_csv(cleaned_data, "data/analysis_data/analysis_data.csv")

### 2024-09-26 scripts/02-data_cleaning.R
cleaned data further and added some tests for data validty

@@ -17,19 +17,35 @@ raw_data <- read_csv("data/raw_data/raw_data.csv")
 filtered <- subset(raw_data, CAPACITY_TYPE == 'Bed Based Capacity' &
                      PROGRAM_MODEL == 'Emergency' & LOCATION_CITY == 'Toronto')
 
+# filters out negative numbers in numeric columns
+filtered <- filtered[filtered$UNOCCUPIED_BEDS >=0, ]
+filtered <- filtered[filtered$OCCUPIED_BEDS >=0, ]
+filtered <- filtered[filtered$CAPACITY_ACTUAL_BED >=0, ]
+filtered <- filtered[filtered$CAPACITY_FUNDING_BED >=0, ]
+filtered <- filtered[filtered$UNAVAILABLE_BEDS >=0, ]
+filtered <- filtered[filtered$OCCUPANCY_RATE_BEDS >=0, ]
+filtered <- filtered[filtered$OCCUPANCY_RATE_BEDS <= 100, ]
+
+# adds column for month + fsa (first 3 digits of postal code) for later analysis
+filtered$month <- as.numeric(format(as.Date(filtered$OCCUPANCY_DATE), "%m"))
+filtered$fsa <- substr(filtered$LOCATION_POSTAL_CODE, start=1, stop=3)
+
+
 # removes unnecessarily variables
-filtered <- filtered %>% select(OCCUPANCY_DATE, SHELTER_ID, 
-                                LOCATION_POSTAL_CODE, CAPACITY_ACTUAL_BED,
-                                CAPACITY_FUNDING_BED, OCCUPIED_BEDS,
-                                UNOCCUPIED_BEDS, UNAVAILABLE_BEDS, 
-                                OCCUPANCY_RATE_BEDS)
+filtered <- filtered %>% select(month, OCCUPANCY_DATE, fsa, LOCATION_ID, 
+                                CAPACITY_ACTUAL_BED, CAPACITY_FUNDING_BED, 
+                                OCCUPIED_BEDS, UNOCCUPIED_BEDS, 
+                                UNAVAILABLE_BEDS, OCCUPANCY_RATE_BEDS)
 
-# adds column for month
-# filtered$month <- format(as.Date(filtered$OCCUPANCY_DATE), "%m")
+# removes NA data entries
+filtered <- filtered %>% drop_na()
 
-# removes rows with NA data, cleans variables names
-cleaned_data <- filtered %>% drop_na() |> janitor::clean_names() 
-
[…]

### 2024-09-26 scripts/03-test_data.R

@@ -12,10 +12,27 @@
 library(tidyverse)
 
 #### Test data ####
-data <- read_csv("data/raw_data/simulated.csv")
+data <- read_csv("data/analysis_data/analysis_data.csv")
 
-# Test there are no negative number of marriages
-data$number_of_marriage |> min() <= 0
+# Tests class of columns are what we expect
+data$date |> class() <= "Date"
+data$month |> class() <= "numeric"
+data$actual_capacity |> class() <= "numeric"
+data$funding_capacity |> class() <= "numeric"
+data$occupied_beds |> class() <= "numeric"
+data$unoccupied_beds |> class() <= "numeric"
+data$unavailable_beds |> class() <= "numeric"
+data$occupancy_rate |> class() <= "numeric"
+data$fsa |> class() <= "character"
 
-# Test for NAs
-all(is.na(data$number_of_marriage))
+# Test for absence of negative numbers in capacity_funding_bed
+data$funding_capacity |> min() >= 0
+
+# Test that occupancy_rate is between 0-100
+data$occupancy_rate |> max() <= 100  
+data$occupancy_rate |> max() >= 0
+
+# Tests that dataset assumption of 
+# funding_capacity = actual_capacity + occupied_beds + unoccupied_beds + 
+#                   unavailable_beds holds
+data$funding_capacity

### 2024-09-26 scripts/03-test_data.R
added condition checks for the relation between variables

@@ -33,6 +33,9 @@ data$occupancy_rate |> max() <= 100
 data$occupancy_rate |> max() >= 0
 
 # Tests that dataset assumption of 
-# funding_capacity = actual_capacity + occupied_beds + unoccupied_beds + 
-#                   unavailable_beds holds
-data$funding_capacity
+# funding_capacity = occupied_beds + unoccupied_beds + unavailable_beds holds
+all(data$funding_capacity - data$occupied_beds - data$unoccupied_beds - 
+  data$unavailable_beds == 0)
+
+all(data$funding_capacity == data$actual_capacity + data$unavailable_beds)
+

### 2024-09-26 scripts/03-test_data.R
copied starter code

@@ -1,7 +1,7 @@
 #### Preamble ####
 # Purpose: Tests data
 # Author: Cher Ning-Li
-# Date: 19 September 2024
+# Date: 27 September 2024
 # Contact: cher.ning@mail.utoronto.ca
 # License: MIT
 # Pre-requisites: Must have data downloaded and cleaned

### 2024-09-26 scripts/04-model.R

@@ -1,0 +1,35 @@
+#### Preamble ####
+# Purpose: Models... [...UPDATE THIS...]
+# Author: Cher Ning-Li
+# Date: 27 September 2024
+# Contact: cher.ning@mail.utoronto.ca
+# License: MIT
+# Pre-requisites: have downloaded, cleaned, and tested data validity
+# Any other information needed? [...UPDATE THIS...]
+
+
+#### Workspace setup ####
+library(tidyverse)
+library(rstanarm)
+
+#### Read data ####
+analysis_data <- read_csv("data/analysis_data/analysis_data.csv")
+
+### Model data ####
+first_model <-
+  stan_glm(
+    formula = flying_time ~ length + width,
+    data = analysis_data,
+    family = gaussian(),
+    prior = normal(location = 0, scale = 2.5, autoscale = TRUE),
+    prior_intercept = normal(location = 0, scale = 2.5, autoscale = TRUE),
+    prior_aux = exponential(rate = 1, autoscale = TRUE),
+    seed = 853
+  )
+
+
+#### Save model ####
+saveRDS(
+  first_model,
+  file = "models/first_model.rds"
+)

### 2024-09-26 scripts/05-replications.R

@@ -1,0 +1,16 @@
+#### Preamble ####
+# Purpose: Replicated graphs from... [...UPDATE THIS...]
+# Author: Cher Ning-Li
+# Date: 27 September 2024
+# Contact: cher.ning@mail.utoronto.ca
+# License: MIT
+# Pre-requisites: have downloaded, cleaned, and modeled from the data
+# Any other information needed? [...UPDATE THIS...]
+
+
+#### Workspace setup ####
+library(tidyverse)
+# [...UPDATE THIS...]
+
+#### Load data ####
+# [...ADD CODE HERE...]

### 2024-09-26 paper/paper.qmd
cleaned data further and added some tests for data validty
minor edits + added basic histogram

@@
 ```{r}
-#| label: fig-marriages
-#| fig-cap: Number of Marriages From 2018-2023
+#| label: fig-occupancyrate-vs-month
+#| fig-cap: Shelter Occupancy Rate in 2024 Over the Months
 #| echo: false
 
-data |> 
-  ggplot(aes(x = date, y = marriage_licenses)) + 
-  geom_point()
+#data |> 
+#  ggplot(aes(x = date, y = occupancy_rate)) + 
+#  geom_point()
+
+hist(data$occupied_beds)
 ```
@@
 
-modelsummary::modelsummary(
-  list(
-    "First model" = first_model
-  ),
-  statistic = "mad",
-  fmt = 2
-)
 ```
@@
 
+If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.
+
 ## Second discussion point
@@
 ## Weaknesses and next steps
+
+Weaknesses and next steps should also be included.
 
@@
 
-# Model details {#sec-model-details}
-
-## Posterior predictive check
-
-## Diagnostics
-
 \newpage

### 2024-09-26 scripts/01-download_data.R

@@ -15,6 +15,8 @@ library(tidyverse)
 # package to use write_csv
 library(readr)
 
+# package <- show_package("21c83b32-d5a8-4106-a54f-010dbe49f6f2")
+
 # accesses link directly to download full data file
 all_data <- read_csv("https://ckan0.cf.opendata.inter.prod-toronto.ca/dataset/21c83b32-d5a8-4106-a54f-010dbe49f6f2/resource/ffd20867-6e3c-4074-8427-d63810edf231/download/Daily%20shelter%20overnight%20occupancy.csv")
 

### 2024-09-26 scripts/03-test_data.R

@@ -15,7 +15,8 @@ library(tidyverse)
 data <- read_csv("data/analysis_data/analysis_data.csv")
 
 # Tests class of columns are what we expect
-data$date |> class() <= "Date"
+expect_equal(class(data$date), "Date") 
+#data$date |> class() <= "Date"
 data$month |> class() <= "numeric"
 data$actual_capacity |> class() <= "numeric"
 data$funding_capacity |> class() <= "numeric"

### 2024-09-26 scripts/04-model.R

@@ -16,6 +16,8 @@ library(rstanarm)
 analysis_data <- read_csv("data/analysis_data/analysis_data.csv")
 
 ### Model data ####
+
+# start with basic histogram 
 first_model <-
   stan_glm(
     formula = flying_time ~ length + width,

### 2024-09-26 paper/paper.qmd
added summary table of means

@@
 title: "Daily Shelter and Overnight Services Availabilities"
-#subtitle: "Marriage Data Cleaning"
 author: "Cher Ning-Li"
@@
 
-Some of our data is of marriages (@fig-marriages), from @opendatatoronto.
+Some of our data is of occupancy rate of shelters across the months (@fig-occupancyrate-vs-month), from @opendatatoronto.
 
+#switch to table here
 ```{r}
@@
 
-#data |> 
-#  ggplot(aes(x = date, y = occupancy_rate)) + 
-#  geom_point()
+rate_means <- data %>% select(month, occupancy_rate) %>% group_by(month) %>% 
+  summarise(mean_occupancy_rate = mean(occupancy_rate))
 
-hist(data$occupied_beds)
+knitr::kable(rate_means, align ="c", col.names= c("Month", "Mean Occupancy Rate"))
+
+```
+# add summary table of funding capacity here
+```{r}
+#| label: fig-fundingcapacity-summary
+#| fig-cap: Funding Capacities of Toronto Emergency Shelters
+#| echo: false
+
+summary(data$funding_capacity)
+```
+
+```{r}
+#| label: fig-fundingcapacity-histogram
+#| fig-cap: Funding Capacities of Toronto Emergency Shelters
+#| echo: false
+
+hist(data$funding_capacity)
 ```
@@
 We can see that the number of marriages fluctuates across the years.
-
-# Model
-
-## Model set-up
-
-We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.
-
-### Model justification
 

### 2024-09-26 paper/paper.qmd
added many hists and diagrams

@@
 
-#switch to table here
+# check if occupancy rate varies over months
 ```{r}
@@
 ```
-# add summary table of funding capacity here
+
+# first find num shelters per fsa
 ```{r}
-#| label: fig-fundingcapacity-summary
-#| fig-cap: Funding Capacities of Toronto Emergency Shelters
+#| label: fig-fsa-numshelters
+#| eval: true
+#| tbl-cap: "Number of Unique Shelters vs FSA"
 #| echo: false
 
-summary(data$funding_capacity)
+fsa_shelters <- data %>% select(fsa, location_id) %>% group_by(fsa) %>% 
+  summarise(numshelters = n_distinct(location_id))
+
+fsa_shelters <- fsa_shelters[order(fsa_shelters$numshelters, decreasing = TRUE), ]
+knitr::kable(top_n(fsa_shelters,5,numshelters), align ="c", col.names= c("FSA", "Number of Shelter Locations"))
+
+```
+```{r}
+hist(fsa_shelters$numshelters)
 ```
@@
 ```{r}
+#| label: fig-regionoccupancy
+#| fig-cap: Average Bed Occupancy of Shelters Across Different FSA Regions
+#| echo: false
+
+region_occupancy <- data %>% select(fsa, occupied_beds) %>% group_by(fsa) %>%
+  summarise(mean_occupied = mean(occupied_beds))
+
+hist(region_occupancy$mean_occupied)
+
+```
+```{r}
 #| label: fig-fundingcapacity-histogram
-#| fig-cap: Funding Capacities of Toronto Emergency Shelters
+#| fig-cap: "Funding Capacities of Toronto Emergency Shelters"
 #| echo: false
@@
 
-We can see that the number of marriages fluctuates across the years.
+# discuss summary stats of funding capacity in text, can consider to only include top few rows of this table
+```{r}
+#| label: fig-fsa-fundingcapacity
+#| eval: true
+#| tbl-cap: "Funding Capacities of Toronto Emergency Shelters"
+#| echo: false
+
+funding_means <- data %>% select(fsa, funding_capacity) %>% group_by(fsa) %>% 
+  summarise(mean_fundingcap = mean(funding_capacity))
+
+funding_means <- funding_means[order(funding_means$mean_fundingcap, decreasing = TRUE), ]
+
+#knitr::kable(top_n(funding_means,5,mean_fundingcap), align ="c", col.names= c("FSA", "Mean Funding Capacity"))
+hist(funding_means$mean_fundingcap)
+```
+
+```{r}
+#| label: fig-unavailabilities-by-fsa
+#| eval: true
+#| tbl-cap: "Number of Unavailable Beds/Night At Different FSAs"
+#| echo: false
+
+unavailable_beds <- data %>% select(fsa, unavailable_beds) %>% group_by(fsa) %>% 
+  summarise(mean_unavailable = mean(unavailable_beds))
+
+unavailable_beds <- unavailable_beds[order(unavailable_beds$mean_unavailable, decreasing = TRUE), ]
+
+knitr::kable(top_n(unavailable_beds,5,unavailable_beds$mean_unavailable), 
+             align ="c", col.names= c("FSA", "Average Number of Unavailable Beds"))
+#hist(unavailable_beds$mean_unavailable)
+```
 

### 2024-09-26 paper/paper.qmd
added variable to simulated data + other minor changes

@@
 ---
-title: "Daily Shelter and Overnight Services Availabilities"
+title: "Emergency Homeless Shelter Availabilities Across Toronto"
 author: "Cher Ning-Li"
@@
 date-format: long
-abstract: "In this paper I analyze the capacity vs real occupancy of shelters across Canada."
+abstract: "In this paper I analyze the occupancy and capacity of emergency homeless shelters across Toronto."
 #TODO Complete abstract after paper.
@@
 
+Next, the @sec-data section will introduce the dataset used for the present analysis and the data cleaning methods applied.
+The following @sec-results will then go more into depth to discover trends and patterns in the data while @sec-discussion will connect these findings to current literature to understand the scope and urgency of the issues at hand.
+
+# Data {#sec-data}
 You can and should cross-reference sections and sub-sections. We use @citeR and @citeTidyverse
 
-The remainder of this paper is structured as follows. @sec-data....
-
-# Data {#sec-data}
-
-Some of our data is of occupancy rate of shelters across the months (@fig-occupancyrate-vs-month), from @opendatatoronto.
+The dataset used for analysis is of occupancy rate of shelters in the year 2024 (@fig-occupancyrate-vs-month), from @opendatatoronto.
 
@@
 #| label: fig-occupancyrate-vs-month
-#| fig-cap: Shelter Occupancy Rate in 2024 Over the Months
+#| tbl-cap: Shelter Occupancy Rate in 2024 Over the Months
 #| echo: false
@@
   summarise(mean_occupancy_rate = mean(occupancy_rate))
+
+rate_means$month <- month.abb[rate_means$month]
 
@@
 
-# Results
+# Results {#sec-results}
 
@@
 
-```{r}
-#| echo: false
-#| eval: true
-#| label: tbl-modelresults
-#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
-#| warning: false
-
-```
-
-# Discussion
+# Discussion {#sec-discussion}
 

### 2024-09-26 scripts/00-simulate_data.R

@@ -15,8 +15,8 @@ library(tidyverse)
 set.seed(304)
 
 # Define the start and end date
-start_date <- as.Date("2018-01-01")
-end_date <- as.Date("2023-12-31")
+start_date <- as.Date("2024-01-01")
+end_date <- as.Date("2024-09-26")
 
 # Set the number of random dates you want to generate
 number_of_dates <- 100
@@ -32,7 +32,8 @@ data <-
       origin = "1970-01-01"
     ),
     capacity = rpois(n = number_of_dates, lambda = 100),
-    vacancies = rpois(n = number_of_dates, lambda = 10)
+    occupied = rpois(n = number_of_dates, lambda = 10),
+    unoccupied = rpois(n = number_of_dates, lambda = 10)
   )
 
 # Write csv

### 2024-09-26 scripts/03-test_data.R

@@ -15,8 +15,7 @@ library(tidyverse)
 data <- read_csv("data/analysis_data/analysis_data.csv")
 
 # Tests class of columns are what we expect
-expect_equal(class(data$date), "Date") 
-#data$date |> class() <= "Date"
+data$date |> class() <= "Date"
 data$month |> class() <= "numeric"
 data$actual_capacity |> class() <= "numeric"
 data$funding_capacity |> class() <= "numeric"

### 2024-09-26 paper/paper.qmd
updated simulated data to add more variables

@@
 
-Next, the @sec-data section will introduce the dataset used for the present analysis and the data cleaning methods applied.
-The following @sec-results will then go more into depth to discover trends and patterns in the data while @sec-discussion will connect these findings to current literature to understand the scope and urgency of the issues at hand.
+Much like many other large cities around the world, Toronto is facing a homelessness crisis that seems to be ever-increasing.
+Whether it is because of environmental factors such as rising housing prices and lack of employment opportunities, or other issues such as drug abuse and mental health struggles, the homeless population of Toronto has risen to an alarming 10,811 in May 2023 (#cite CBC).
+The city council has recognized the urgency of this problem, declaring homelessness an emergency, and pledged their full efforts towards helping the homeless population move into safe homes (#CBC).
+Since then, the city of Toronto has focussed on securing more financial support from the provincial and federal government, requesting up to $25 million to fund various housing projects.
+Notably, the Canada-Ontario Housing Benefit (COHB) aimed to move 1,600 — 2,000 households access and transition into permanent housing in 2023-2024 (#CBC).
+
+So, how did these efforts turn out?
+Analyzing Toronto's shelter occupancy data from 2024, we do not see optimistic results.
+Average occupancy rate of emergency shelters per night remains at 99%, with no significant decrease across the months (#cite dataset).
+Despite the evident high demand, there is even a decrease, albeit small, in shelters' funding capacity over the months (#cite dataset).
+
+In this paper, emergency shelter occupancy and availability will be compared across different months as well as different regions of Toronto to search for patterns or fluctutations in demand.
+The region of each shelter will be determined by its forward sorting area (FSA) code, or the first three digits of its postal code.
+As well, the regions that observe the highest frequency of failing to make efficient use of funding are identified as potential problem areas; further investigation into potential causes would be beneficial to understand if there are improvements that can be made so that the limited available funding can be maximized.
+
+To achieve this, the @sec-data section will introduce the Toronto Shelter & Support Services dataset used for the present analysis and the data cleaning methods applied.
+The following @sec-discussion will then go more into depth to apply context to the patterns present in the data, aiming to understand the scope and urgency of the issues at hand.
 
@@
 
-The dataset used for analysis is of occupancy rate of shelters in the year 2024 (@fig-occupancyrate-vs-month), from @opendatatoronto.
+The dataset used for analysis is of occupancy rate of shelters in the year 2024 (@tbl-occupancyrate-vs-month), from @opendatatoronto.
 
@@
 ```{r}
-#| label: fig-occupancyrate-vs-month
+#| label: tbl-occupancyrate-vs-month
 #| tbl-cap: Shelter Occupancy Rate in 2024 Over the Months
@@
 
-rate_means <- data %>% select(month, occupancy_rate) %>% group_by(month) %>% 
-  summarise(mean_occupancy_rate = mean(occupancy_rate))
+rate_means <- data %>% select(month, occupancy_rate, funding_capacity) %>% group_by(month) %>% 
+  summarise(mean_occupancy_rate = mean(occupancy_rate), mean_capacity = mean(funding_capacity))
 
@@
 
-knitr::kable(rate_means, align ="c", col.names= c("Month", "Mean Occupancy Rate"))
+knitr::kable(rate_means, align ="c", col.names= c("Month", "Mean Occupancy Rate", "Mean Funding Capacity"))
 
@@
 ```{r}
+#| label: fig-numshelters-vs-fsa
+#| fig-cap: Number of Shelters in Each FSA
+#| echo: false
+
 hist(fsa_shelters$numshelters)
@@
 # Discussion {#sec-discussion}
+Additionally, the distribution of shelter locations across the city's different forward sortation areas (FSA), or the first three digits of a zip code, displays that there are particular areas with higher demand than others.
+For example, particular blocks in Old Toronto are evidently areas that require higher attention, as they are among the FSAs to receive most funding as well as the ones that have highest number of shelters (#cite dataset).
 
 ## First discussion point {#sec-first-point}
-
-If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.
 

### 2024-09-26 scripts/00-simulate_data.R

@@ -21,7 +21,7 @@ end_date <- as.Date("2024-09-26")
 # Set the number of random dates you want to generate
 number_of_dates <- 100
 
-data <-
+simulated_data <-
   tibble(
     dates = as.Date(
       runif(
@@ -31,10 +31,11 @@ data <-
       ),
       origin = "1970-01-01"
     ),
-    capacity = rpois(n = number_of_dates, lambda = 100),
-    occupied = rpois(n = number_of_dates, lambda = 10),
-    unoccupied = rpois(n = number_of_dates, lambda = 10)
+    capacity = rpois(n = number_of_dates, lambda = 50),
+    occupied = min(capacity, rpois(n = number_of_dates, lambda = 50)),
+    unoccupied = capacity - occupied,
+    occupancy_rate = (occupied/capacity) *100
   )
 
 # Write csv
-write_csv(data, file = "data/raw_data/simulated.csv")
+write_csv(simulated_data, file = "data/raw_data/simulated.csv")

### 2024-09-27 paper/paper.qmd
completed data overview + introduction

@@
 number-sections: true
+toc: true
 bibliography: references.bib
@@
 Since then, the city of Toronto has focussed on securing more financial support from the provincial and federal government, requesting up to $25 million to fund various housing projects.
-Notably, the Canada-Ontario Housing Benefit (COHB) aimed to move 1,600 — 2,000 households access and transition into permanent housing in 2023-2024 (#CBC).
+Notably, the Canada-Ontario Housing Benefit (COHB) aimed to move 1,600-2,000 households access and transition into permanent housing in 2023-2024 (#CBC).
 
@@
 
-To achieve this, the @sec-data section will introduce the Toronto Shelter & Support Services dataset used for the present analysis and the data cleaning methods applied.
+To achieve this, @sec-data will introduce the Toronto Shelter & Support Services dataset used for the present analysis and the data cleaning methods applied.
 The following @sec-discussion will then go more into depth to apply context to the patterns present in the data, aiming to understand the scope and urgency of the issues at hand.
@@
 # Data {#sec-data}
-You can and should cross-reference sections and sub-sections. We use @citeR and @citeTidyverse
 
-The dataset used for analysis is of occupancy rate of shelters in the year 2024 (@tbl-occupancyrate-vs-month), from @opendatatoronto.
+## Overview {#data-overview}
 
-# check if occupancy rate varies over months
+The dataset used for analysis is of shelter occupancy in the year 2024 (#cite dataset), from @opendatatoronto.
+The data is updated daily by the Toronto Shelter & Support Services and has the Open Data License @tphlicense.
+However, one limitation it faces is that it is unaudited, therefore there are limited verification methods to ensure that the shelter programs' records are accurate to their actual situation.
+
+All shelters in this dataset are classified as either Emergency or Transitional programs, with Transitional locations providing more specialized programming and being exclusively offered to eligible individuals (#cite dataset).
+With consideration for the greater accessibility of Emergency programs to the entire homeless population as well as the greater number of them shelters compared to Transitional, this analysis will focus on Emergency programs only.
+As well, all recorded shelters either measure capacity based on number of beds or rooms; similarly, this analysis will focus only on bed-based capacity shelters due to its greater prevalence.
+This means that all entries pertaining to non-Toronto based, Transitional type, or room-based capacity shelters were not considered in this analysis.
+
+With the remaining data entries, the variables of interest were the shelter's location, funding capacity, actual capacity, occupied beds, unoccupied beds, unavailable beds, and occupancy rate.
+Funding capacity refers to the maximum number of beds the location is able to offer, and represents the sum of actual capacity and unavailable beds.
+Unavailable beds refers to the number of spaces that are out of service due to logistical reasons (#cite dataset).
+Actual capacity represents the number of beds that are usable per night, and should be the sum of occupied and unoccupied beds.
+After the dataset was cleaned for entries with empty values and invalid negative values, all entries in the remaining set were shown to adhere to these relations when tested.
+
+The R Programming language [@citeR] was used for all data cleaning, testing, and analysis.
+The packages `opendatatoronto` [@opendatatoronto], `tidyverse` [@citeTidyverse], `dplyr` [@citeDplyr], and `readr` [@citeReadr] were used to simulate and download data.
+Packages `tidyverse` [@citeTidyverse] and `janitor` [@citeJanitor] were then used to clean and test the downloaded dataset.
+
+## Results {#data-findings}
+
+(# check if occupancy rate varies over months
 ```{r}
@@
 
-# first find num shelters per fsa
+(#first find num shelters per fsa
 ```{r}
@@
 
-# discuss summary stats of funding capacity in text, can consider to only include top few rows of this table
+(#discuss summary stats of funding capacity in text, can consider to only include top few rows of this table
 ```{r}
@@
 
-# Results {#sec-results}
-
-Our results are summarized in @tbl-modelresults.
-
-```{r}
-#| echo: false
-#| eval: true
-#| warning: false
-#| message: false
-
-library(rstanarm)
-
-```
-
 # Discussion {#sec-discussion}
@@
 # Appendix {-}
-
+## Data License Information
+Contains information licensed under the Open Government Licence – Toronto.
 # Additional data details

### 2024-09-27 scripts/04-model.R

@@ -1,37 +1,0 @@
-#### Preamble ####
-# Purpose: Models... [...UPDATE THIS...]
-# Author: Cher Ning-Li
-# Date: 27 September 2024
-# Contact: cher.ning@mail.utoronto.ca
-# License: MIT
-# Pre-requisites: have downloaded, cleaned, and tested data validity
-# Any other information needed? [...UPDATE THIS...]
-
-
-#### Workspace setup ####
-library(tidyverse)
-library(rstanarm)
-
-#### Read data ####
-analysis_data <- read_csv("data/analysis_data/analysis_data.csv")
-
-### Model data ####
-
-# start with basic histogram 
-first_model <-
-  stan_glm(
-    formula = flying_time ~ length + width,
-    data = analysis_data,
-    family = gaussian(),
-    prior = normal(location = 0, scale = 2.5, autoscale = TRUE),
-    prior_intercept = normal(location = 0, scale = 2.5, autoscale = TRUE),
-    prior_aux = exponential(rate = 1, autoscale = TRUE),
-    seed = 853
-  )
-
-
-#### Save model ####
-saveRDS(
-  first_model,
-  file = "models/first_model.rds"
-)

### 2024-09-27 scripts/05-replications.R

@@ -1,16 +1,0 @@
-#### Preamble ####
-# Purpose: Replicated graphs from... [...UPDATE THIS...]
-# Author: Cher Ning-Li
-# Date: 27 September 2024
-# Contact: cher.ning@mail.utoronto.ca
-# License: MIT
-# Pre-requisites: have downloaded, cleaned, and modeled from the data
-# Any other information needed? [...UPDATE THIS...]
-
-
-#### Workspace setup ####
-library(tidyverse)
-# [...UPDATE THIS...]
-
-#### Load data ####
-# [...ADD CODE HERE...]

### 2024-09-27 paper/paper.qmd
fixed some citations

@@
 Much like many other large cities around the world, Toronto is facing a homelessness crisis that seems to be ever-increasing.
-Whether it is because of environmental factors such as rising housing prices and lack of employment opportunities, or other issues such as drug abuse and mental health struggles, the homeless population of Toronto has risen to an alarming 10,811 in May 2023 (#cite CBC).
-The city council has recognized the urgency of this problem, declaring homelessness an emergency, and pledged their full efforts towards helping the homeless population move into safe homes (#CBC).
+Whether it is because of environmental factors such as rising housing prices and lack of employment opportunities, or other issues such as drug abuse and mental health struggles, the homeless population of Toronto has risen to an alarming 10,811 in May 2023 [@cbcArticle].
+The city council has recognized the urgency of this problem, declaring homelessness an emergency, and pledged their full efforts towards helping the homeless population move into safe homes [@cbcArticle].
 Since then, the city of Toronto has focussed on securing more financial support from the provincial and federal government, requesting up to $25 million to fund various housing projects.
-Notably, the Canada-Ontario Housing Benefit (COHB) aimed to move 1,600-2,000 households access and transition into permanent housing in 2023-2024 (#CBC).
+Notably, the Canada-Ontario Housing Benefit (COHB) aimed to move 1,600-2,000 households access and transition into permanent housing in 2023-2024 [@cbcArticle].
 
@@
 Analyzing Toronto's shelter occupancy data from 2024, we do not see optimistic results.
-Average occupancy rate of emergency shelters per night remains at 99%, with no significant decrease across the months (#cite dataset).
-Despite the evident high demand, there is even a decrease, albeit small, in shelters' funding capacity over the months (#cite dataset).
+Average occupancy rate of emergency shelters per night remains at 99%, with no significant decrease across the months [@shelterDataset].
+Despite the evident high demand, there is even a decrease, albeit small, in shelters' funding capacity over the months [@shelterDataset].
 
@@
 
-The dataset used for analysis is of shelter occupancy in the year 2024 (#cite dataset), from @opendatatoronto.
-The data is updated daily by the Toronto Shelter & Support Services and has the Open Data License @tphlicense.
+The dataset used for analysis is of shelter occupancy in the year 2024 [@shelterDataset], from Open Data Toronto [@opendatatoronto].
+The data is updated daily by the Toronto Shelter & Support Services and has the Open Data License (@sec-license).
 However, one limitation it faces is that it is unaudited, therefore there are limited verification methods to ensure that the shelter programs' records are accurate to their actual situation.
 
-All shelters in this dataset are classified as either Emergency or Transitional programs, with Transitional locations providing more specialized programming and being exclusively offered to eligible individuals (#cite dataset).
-With consideration for the greater accessibility of Emergency programs to the entire homeless population as well as the greater number of them shelters compared to Transitional, this analysis will focus on Emergency programs only.
-As well, all recorded shelters either measure capacity based on number of beds or rooms; similarly, this analysis will focus only on bed-based capacity shelters due to its greater prevalence.
+All shelters in this dataset are classified as either Emergency or Transitional programs, with Transitional locations providing more specialized programming and being exclusively offered to eligible individuals [@shelterDataset].
+With consideration for the greater accessibility of Emergency programs to the entire homeless population as well as the greater number of them shelters compared to Transitional, this analysis will focus on Emergency programs only [@shelterDataset].
+As well, all recorded shelters either measure capacity based on number of beds or rooms; similarly, this analysis will focus only on bed-based capacity shelters due to its greater prevalence [@shelterDataset].
 This means that all entries pertaining to non-Toronto based, Transitional type, or room-based capacity shelters were not considered in this analysis.
@@
 Funding capacity refers to the maximum number of beds the location is able to offer, and represents the sum of actual capacity and unavailable beds.
-Unavailable beds refers to the number of spaces that are out of service due to logistical reasons (#cite dataset).
+Unavailable beds refers to the number of spaces that are out of service due to logistical reasons [@shelterDataset].
 Actual capacity represents the number of beds that are usable per night, and should be the sum of occupied and unoccupied beds.
@@
 
-# Appendix {-}
-## Data License Information
-Contains information licensed under the Open Government Licence – Toronto.
+# Appendix {#sec-appendix}
+## Data License Information {#sec-license}
+"Contains information licensed under the Open Government Licence – Toronto" [@tphlicense].
 # Additional data details

### 2024-09-27 paper/paper.qmd
added more histograms

@@
 author: "Cher Ning-Li"
-thanks: "Code and data are available at: https://github.com/cher-ning/shelter_occupancies"
-date: today
+thanks: "Code and data are available at: <https://github.com/cher-ning/shelter_occupancies>"
+date: 09-27-2024
 date-format: long
@@
 library(tidyverse)
+library(ggplot2)
 
@@
 This means that all entries pertaining to non-Toronto based, Transitional type, or room-based capacity shelters were not considered in this analysis.
+It should also be noted that as this dataset contains only data from this year, only data from January 1 to September 26 is accessible.
 
@@
 ## Results {#data-findings}
+This following section will use packages `ggplot2` [@citeggplot] and `knitr` [@citeKnitr] to generate several plots and tables to assist with visualizing patterns in the dataset.
 
-(# check if occupancy rate varies over months
+@tbl-occupancyrate-vs-month provides an initial summary of the shelters' mean occupancy rates over the months.
+As observed, there is consistent and high demand for emergency shelters regardless of the time of year.
+
 ```{r}
 #| label: tbl-occupancyrate-vs-month
-#| tbl-cap: Shelter Occupancy Rate in 2024 Over the Months
+#| tbl-cap: 2024 Shelter Occupancy Rate Over the Months
 #| echo: false
 
-rate_means <- data %>% select(month, occupancy_rate, funding_capacity) %>% group_by(month) %>% 
-  summarise(mean_occupancy_rate = mean(occupancy_rate), mean_capacity = mean(funding_capacity))
+# checks if occupancy rate varies over months
+
+rate_means <- data %>% select(month, occupancy_rate) %>% group_by(month) %>% 
+  summarise(mean_occupancy_rate = round(mean(occupancy_rate), 2))
 
 rate_means$month <- month.abb[rate_means$month]
-
-knitr::kable(rate_means, align ="c", col.names= c("Month", "Mean Occupancy Rate", "Mean Funding Capacity"))
+knitr::kable(rate_means, align ="c", col.names= c("Month", "Occupancy Rate"))
 
@@
 
-(#first find num shelters per fsa
+The occupancy rate can then be broken down into the mean funding capacity, actual capacity, and real number of occupied beds per month.
+The high occupancy rate observed in @tbl-occupancyrate-vs-month makes it unsurprising to see the number of occupied beds consistently match so closely to the shelters' actual capacity.
+Notably, there is a significant decrease in funding around March, which will be further discussed in @sec-discussion.
+It is also important to note that despite the near 100% occupancy rate, there is often a gap between the funding capacity and actual capacity, suggesting that sometimes the given facilities may not be utilized at their maximum potential.
+
 ```{r}
-#| label: fig-fsa-numshelters
+#| label: fig-occupancy-vs-month
+#| fig-cap: Shelter Capacity and Occupancy From Jan to Sep 2024
+#| echo: false
+
+# checks occupancy and capacity over months
+occupancy_means <- data %>% select(month, occupied_beds, funding_capacity, actual_capacity) %>% group_by(month) %>% 
+  summarise(mean_occupied = mean(occupied_beds), mean_capacity = mean(funding_capacity), mean_actual = mean(actual_capacity))
+
+ggplot(occupancy_means, aes(x=month)) + 
+  geom_line(aes(y = mean_occupied, colour = "Occupied Beds")) + 
+  geom_line(aes(y = mean_capacity, colour="Funding Capacity")) +
+  geom_line(aes(y = mean_actual, colour="Actual Capacity")) + 
+  scale_colour_manual("", 
+                      values = c("Occupied Beds"="darkred", "Funding Capacity"="steelblue", 
+                                 "Actual Capacity"="darkblue")) +
+  xlab(" ") +
+  theme_classic() +
+  labs(x = "Month", y = "Number of Beds") + 
+  scale_x_continuous(breaks = seq(1, 9, by = 1))
+
+```
+
+Another dimension to consider is the geographical region of the shelters.
+Still focussing on the mean number of occupied beds, @fig-regionoccupancy shows a rather wide spread across different FSA regions, ranging from a minimum of 18.60 to a maximum of 87.74.
+Some of the regions reporting on the higher end include M5H, M6P, and M2H, which are neighbourhoods in Old Toronto, North York, and around High Park [@citeFSA].
+
+```{r}
+#| label: fig-regionoccupancy
+#| fig-cap: Average Bed Occupancy of Shelters in Each FSA Region
+#| echo: false
+
+region_occupancy <- data %>% select(fsa, occupied_beds) %>% group_by(fsa) %>%
+  summarise(mean_occupied = mean(occupied_beds))
+
+ggplot(region_occupancy, aes(x=mean_occupied)) + geom_histogram(binwidth = 7) +
+  labs(x = "Mean Occupancy in Each FSA", y = "Count") + theme_classic()
+```
+
+```{r}
+#| echo: false
+#| include: false
+
+ordered_region_occupancy <- region_occupancy[order(region_occupancy$mean_occupied, decreasing = TRUE), ]
+ordered_region_occupancy
+```
+
+Another dimension to consider in analyzing which regions may be experiencing higher rates of homelessness is the number of distinct shelter locations made available there.
+Sorting for unique shelter ID's across the 32 FSA regions in this data set, @tbl-fsa-numshelters and @fig-numshelters-vs-fsa show that regions most commonly have only 1-2 shelters each, but there are areas with particularly high density.
+Specifically, the region M5A located in Old Toronto [@citeFSA] has 7 shelters, which is several times the number placed in other regions.
+
+```{r}
+#| label: tbl-fsa-numshelters
 #| eval: true
-#| tbl-cap: "Number of Unique Shelters vs FSA"
+#| tbl-cap: "Number of Emergency Shelter Locations Reported in Each Toronto FSA"
+#| echo: false
+
+#first find num shelters per fsa
+
+fsa_shelters <- data %>% select(fsa, location_id) %>% group_by(fsa) %>% 
+  summarise(numshelters = n_distinct(location_id))
+fsa_shelters <- fsa_shelters[order(fsa_shelters$numshelters, decreasing = TRUE), ]
+knitr::kable(top_n(fsa_shelters, 4, numshelters), align ="c", col.names= c("FSA", "Number of Shelter Locations"))
+```
+
+```{r}
+#| label: fig-numshelters-vs-fsa
+#| fig-cap: Number of Shelters in Each FSA
 #| echo: false
@@
 
-fsa_shelters <- fsa_shelters[order(fsa_shelters$numshelters, decreasing = TRUE), ]
-knitr::kable(top_n(fsa_shelters,5,numshelters), align ="c", col.names= c("FSA", "Number of Shelter Locations"))
-
-```
-```{r}
-#| label: fig-numshelters-vs-fsa
-#| fig-cap: Number of Shelters in Each FSA
-#| echo: false
-
-hist(fsa_shelters$numshelters)
+ggplot(fsa_shelters, aes(x=numshelters)) + geom_histogram(binwidth = 1) +
+  labs(x = "Number of Shelters in Each FSA", y = "Count") + theme_classic()
 ```
 
-```{r}
-#| label: fig-regionoccupancy
-#| fig-cap: Average Bed Occupancy of Shelters Across Different FSA Regions
-#| echo: false
-
-region_occupancy <- data %>% select(fsa, occupied_beds) %>% group_by(fsa) %>%
-  summarise(mean_occupied = mean(occupied_beds))
-
-hist(region_occupancy$mean_occupied)
-
-```
 ```{r}
@@
 
-(#discuss summary stats of funding capacity in text, can consider to only include top few rows of this table
 ```{r}

### 2024-09-27 paper/paper.qmd
added literature citation + finished up data section

@@
 
+#install.packages("tidyverse")
+#install.packages("ggplot2")
+#install.packages("knitr")
 library(tidyverse)
 library(ggplot2)
+library(knitr)
 
@@
 
+\newpage
 # Introduction
@@
 
-So, how did these efforts turn out?
+In addition to these long-term plans that aim to move homeless individuals into stable homes, temporary shelters are also crucial during this transition process to provide short-term safety and resources.
+Research has shown that without access to shelter services, individuals can be pushed into substance-use relapses and hospitalizations, food insecurity, institutional circuitry through criminal justice systems, and other factors of instability [@citeKerman].
+Shelters also provide useful data to track and understand the current state of the city's battle against this crisis.
+
+With that in mind, how did the city council's efforts turn out?
 Analyzing Toronto's shelter occupancy data from 2024, we do not see optimistic results.
@@
 The following @sec-discussion will then go more into depth to apply context to the patterns present in the data, aiming to understand the scope and urgency of the issues at hand.
+
+\newpage
 
@@
 This means that all entries pertaining to non-Toronto based, Transitional type, or room-based capacity shelters were not considered in this analysis.
-It should also be noted that as this dataset contains only data from this year, only data from January 1 to September 26 is accessible.
+It should also be noted that as this dataset contains only data from this year, only data from January 1 to September 26 was accessible as of writing this paper, and therefore only these entries were utilized in analysis.
 
@@
 Packages `tidyverse` [@citeTidyverse] and `janitor` [@citeJanitor] were then used to clean and test the downloaded dataset.
+
+\newpage
 
@@
 The occupancy rate can then be broken down into the mean funding capacity, actual capacity, and real number of occupied beds per month.
-The high occupancy rate observed in @tbl-occupancyrate-vs-month makes it unsurprising to see the number of occupied beds consistently match so closely to the shelters' actual capacity.
-Notably, there is a significant decrease in funding around March, which will be further discussed in @sec-discussion.
+The high occupancy rate observed in @tbl-occupancyrate-vs-month makes it unsurprising to see the number of occupied beds consistently match so closely to the shelters' actual capacity in @fig-occupancy-vs-month.
+Notably, @fig-occupancy-vs-month also shows a significant decrease in funding around March, which will be further discussed in @sec-discussion.
 It is also important to note that despite the near 100% occupancy rate, there is often a gap between the funding capacity and actual capacity, suggesting that sometimes the given facilities may not be utilized at their maximum potential.
-
 ```{r}
@@
 ```
-
 Another dimension to consider is the geographical region of the shelters.
@@
 Some of the regions reporting on the higher end include M5H, M6P, and M2H, which are neighbourhoods in Old Toronto, North York, and around High Park [@citeFSA].
-
 ```{r}
@@
 #| include: false
+#| eval: true
 
 ordered_region_occupancy <- region_occupancy[order(region_occupancy$mean_occupied, decreasing = TRUE), ]
-ordered_region_occupancy
 ```
@@
 
-```{r}
-#| label: fig-fundingcapacity-histogram
-#| fig-cap: "Funding Capacities of Toronto Emergency Shelters"
-#| echo: false
+As shown in @fig-occupancy-vs-month, there is often a gap between a shelter's funding capacity and its actual capacity.
+This difference, reported as the number of unavailable beds in this dataset, represents resources that should have been available to the homeless population but was not on a particular night, either due to maintenance, renovations, outbreaks, or pest control issues [@shelterDataset].
 
-hist(data$funding_capacity)
-```
+Using similar methods as above to group data by FSA code finds that the average number of unavailable bed spaces per region is 1.30.
+Despite this promisingly low number, a closer look at the distribution in @tbl-unavailabilities-by-fsa reveals that there is a significantly higher average of unavailability in M2N and M5G, representing neighbourhoods in North York and Old Toronto respectively [@citeFSA].
+The implications of this finding will be discussed further in @sec-discussion.
 
 ```{r}
-#| label: fig-fsa-fundingcapacity
-#| eval: true
-#| tbl-cap: "Funding Capacities of Toronto Emergency Shelters"
-#| echo: false
-
-funding_means <- data %>% select(fsa, funding_capacity) %>% group_by(fsa) %>% 
-  summarise(mean_fundingcap = mean(funding_capacity))
-
-funding_means <- funding_means[order(funding_means$mean_fundingcap, decreasing = TRUE), ]
-
-#knitr::kable(top_n(funding_means,5,mean_fundingcap), align ="c", col.names= c("FSA", "Mean Funding Capacity"))
-hist(funding_means$mean_fundingcap)
-```
-
-```{r}
-#| label: fig-unavailabilities-by-fsa
+#| label: tbl-unavailabilities-by-fsa
 #| eval: true
@@
 unavailable_beds <- data %>% select(fsa, unavailable_beds) %>% group_by(fsa) %>% 
-  summarise(mean_unavailable = mean(unavailable_beds))
+  summarise(mean_unavailable = round(mean(unavailable_beds),3))
 
@@
              align ="c", col.names= c("FSA", "Average Number of Unavailable Beds"))
-#hist(unavailable_beds$mean_unavailable)
+```
+
+```{r}
+#| eval: true
+#| include: false
+#| echo: false
+
+summary(unavailable_beds)
 ```
@@
 # Discussion {#sec-discussion}
+
 Additionally, the distribution of shelter locations across the city's different forward sortation areas (FSA), or the first three digits of a zip code, displays that there are particular areas with higher demand than others.
@@
 
-## First discussion point {#sec-first-point}
+## Main Findings {#sec-discussion-findings}
+Analysis of this dataset in @data-findings yielded several conclusions about the still-prevalent homelessness crisis here in Toronto.
+First, the consistently high occupancy rate throughout the year suggests that the need for housing resources remains high, and is not particularly affected by time of year.
+However, as observed in @fig-occupancy-vs-month, there is a relatively steep dip in funding capacity around March that is not preceded by a dip in demand.
+This decrease is potentially due to the city emphasizing their efforts during winter months, as those are intuitively the most dangerous months to unable to access warmth and shelter [@cbcArticle].
 
-## Second discussion point
+## Implications {#sec-discussion-implications}
+The observations made from this dataset reveal that the city council's effort have seen limited efficacy.
+Occupancy rates at emergency shelters remains alarmingly high all across Toronto with no sign of decreasing.
+Though these short-term solutions
 
-## Third discussion point
-
-## Weaknesses and next steps
-
-Weaknesses and next steps should also be included.
+## Limitations and Next Steps {#sec-discussion-limitations}
 

### 2024-09-27 paper/paper.qmd
added more content to discussion section

@@
 
-## Overview {#data-overview}
+## Overview {#sec-data-overview}
 
@@
 
-## Results {#data-findings}
+## Results {#sec-data-findings}
 This following section will use packages `ggplot2` [@citeggplot] and `knitr` [@citeKnitr] to generate several plots and tables to assist with visualizing patterns in the dataset.
 
-@tbl-occupancyrate-vs-month provides an initial summary of the shelters' mean occupancy rates over the months.
-As observed, there is consistent and high demand for emergency shelters regardless of the time of year.
+@tbl-occupancyrate-vs-month provides a summary of the shelters' mean occupancy rates over the months. As observed, there is consistently high demand for shelters regardless of the time of year.
 
@@
 
-The occupancy rate can then be broken down into the mean funding capacity, actual capacity, and real number of occupied beds per month.
-The high occupancy rate observed in @tbl-occupancyrate-vs-month makes it unsurprising to see the number of occupied beds consistently match so closely to the shelters' actual capacity in @fig-occupancy-vs-month.
-Notably, @fig-occupancy-vs-month also shows a significant decrease in funding around March, which will be further discussed in @sec-discussion.
-It is also important to note that despite the near 100% occupancy rate, there is often a gap between the funding capacity and actual capacity, suggesting that sometimes the given facilities may not be utilized at their maximum potential.
 ```{r}
@@
   scale_x_continuous(breaks = seq(1, 9, by = 1))
+```
 
-```
+\newpage
+The occupancy rate can then be broken down into the mean funding capacity, actual capacity, and real number of occupied beds per month.
+The high occupancy rate observed in @tbl-occupancyrate-vs-month makes it unsurprising to see the number of occupied beds consistently match so closely to the shelters' actual capacity in @fig-occupancy-vs-month.
+Notably, @fig-occupancy-vs-month also shows a significant decrease in funding around March, which will be further discussed in @sec-discussion.
+It is also important to note that despite the near 100% occupancy rate, there is often a gap between the funding capacity and actual capacity, suggesting that sometimes the given facilities may not be utilized at their maximum potential.
+This is a potential area of concern, and its implications will also be discussed in @sec-discussion.
+
 Another dimension to consider is the geographical region of the shelters.
@@
 #| eval: true
-#| tbl-cap: "Number of Unavailable Beds/Night At Different FSAs"
+#| tbl-cap: "Number of Unavailable Beds Per Night in Different FSAs"
 #| echo: false
@@
 
-Additionally, the distribution of shelter locations across the city's different forward sortation areas (FSA), or the first three digits of a zip code, displays that there are particular areas with higher demand than others.
-For example, particular blocks in Old Toronto are evidently areas that require higher attention, as they are among the FSAs to receive most funding as well as the ones that have highest number of shelters (#cite dataset).
+## Main Findings {#sec-discussion-findings}
+Analysis of this dataset in @sec-data-findings yielded several conclusions about the still-prevalent homelessness crisis here in Toronto.
+First, the consistently high occupancy rate throughout the year [@tbl-occupancyrate-vs-month] suggests that the need for housing resources remains high, and is not particularly affected by time of year.
+However, as observed in @fig-occupancy-vs-month, there is a relatively steep dip in funding capacity around March that is not preceded by a noticeable dip in demand.
+This decrease is potentially due to the city emphasizing their efforts during winter months, as those are intuitively the most dangerous months for those living on the streets [@cbcArticle].
+As a result, when winter ends and weather begins warming up in March, the additional funding is no longer available, causing this large and sudden decrease.
 
-## Main Findings {#sec-discussion-findings}
-Analysis of this dataset in @data-findings yielded several conclusions about the still-prevalent homelessness crisis here in Toronto.
-First, the consistently high occupancy rate throughout the year suggests that the need for housing resources remains high, and is not particularly affected by time of year.
-However, as observed in @fig-occupancy-vs-month, there is a relatively steep dip in funding capacity around March that is not preceded by a dip in demand.
-This decrease is potentially due to the city emphasizing their efforts during winter months, as those are intuitively the most dangerous months to unable to access warmth and shelter [@cbcArticle].
+Moving onto FSA-based observations, it was found that there are several neighbourhoods including around Old Toronto and North York that have some of the highest concentration of emergency shelters and have highest funding capacities [@tbl-fsa-numshelters, @fig-numshelters-vs-fsa].
+Despite the high occupancy [@fig-regionoccupancy] in these specific areas, it is important to remember that these findings are purely correlational and not causational.
+From this data alone, it is impossible to conclude whether these areas initially had greater rates of homelessness, thereby leading the government to spend more funds building shelters here, or that it is the higher concentration and availability of shelters here that will attract more homeless population to migrate here.
+
+Next, analysis of the average bed unavailabilities across FSAs
 

### 2024-09-27 paper/paper.qmd
added another cited paper, completed abstract, added to discussion implications

@@
 date-format: long
-abstract: "In this paper I analyze the occupancy and capacity of emergency homeless shelters across Toronto."
-#TODO Complete abstract after paper.
+abstract: "This paper analyzes the occupancy and capacity of emergency homeless shelters across Toronto, noting the high demand across all regions throughout the months.
+Dividing analysis of the data into different neighbourhoods, areas in Old Toronto are identified to have higher bed occupancies and number of shelters.
+Similarly, problem areas in North York and Old Toronto with patterns of inability to provide the number of bed spaces promised are also identified.
+Combining with the context of the current literature researching homelessness, these findings indicate a high urgency for the Toronto city council to implement effective short-term and long-term support programs to target the current homelessness crisis."
 format: pdf
@@
 Packages `tidyverse` [@citeTidyverse] and `janitor` [@citeJanitor] were then used to clean and test the downloaded dataset.
+The R code used to clean, test, and display data were adapted from @citeTB.
 
@@
 
-Another dimension to consider in analyzing which regions may be experiencing higher rates of homelessness is the number of distinct shelter locations made available there.
+Another variable to consider when searching for patterns between geographical region and rates of homelessness is the number of distinct shelter locations made available there.
 Sorting for unique shelter ID's across the 32 FSA regions in this data set, @tbl-fsa-numshelters and @fig-numshelters-vs-fsa show that regions most commonly have only 1-2 shelters each, but there are areas with particularly high density.
@@
 fsa_shelters <- fsa_shelters[order(fsa_shelters$numshelters, decreasing = TRUE), ]
-knitr::kable(top_n(fsa_shelters, 4, numshelters), align ="c", col.names= c("FSA", "Number of Shelter Locations"))
+top_fsa_shelters <- fsa_shelters %>% top_n(3, numshelters)
+knitr::kable(top_fsa_shelters, align ="c", col.names= c("FSA", "Number of Shelter Locations"))
 ```
@@
 
-As shown in @fig-occupancy-vs-month, there is often a gap between a shelter's funding capacity and its actual capacity.
+As shown previously in @fig-occupancy-vs-month, there is often a gap between a shelter's funding capacity and its actual capacity.
 This difference, reported as the number of unavailable beds in this dataset, represents resources that should have been available to the homeless population but was not on a particular night, either due to maintenance, renovations, outbreaks, or pest control issues [@shelterDataset].
 
-Using similar methods as above to group data by FSA code finds that the average number of unavailable bed spaces per region is 1.30.
-Despite this promisingly low number, a closer look at the distribution in @tbl-unavailabilities-by-fsa reveals that there is a significantly higher average of unavailability in M2N and M5G, representing neighbourhoods in North York and Old Toronto respectively [@citeFSA].
+Using similar methods as above to group data by FSA code, the average number of unavailable bed spaces per region is found to be 1.30.
+
+Despite this promisingly low number, @tbl-unavailabilities-by-fsa displays some of the FSAs with highest average unavailability per night and shows that the distribution across regions is noticeably skewed.
+M2N and M5G especially stand out, representing neighbourhoods in North York and Old Toronto respectively [@citeFSA].
+The unavailability as a percentage of funding capacity was also calculated to verify whether this gap could be simply proportional to the total capacity in these regions, and results indicate that this is not the case.
+Even when looking at only the unavailability as a percentage, these two regions are still significantly higher than others.
 The implications of this finding will be discussed further in @sec-discussion.
@@
 #| eval: true
-#| tbl-cap: "Number of Unavailable Beds Per Night in Different FSAs"
+#| tbl-cap: "Unavailable Beds Per Night in Different FSAs"
 #| echo: false
 
-unavailable_beds <- data %>% select(fsa, unavailable_beds) %>% group_by(fsa) %>% 
-  summarise(mean_unavailable = round(mean(unavailable_beds),3))
+unavailable_beds <- data %>% select(fsa, unavailable_beds, funding_capacity) %>% group_by(fsa) %>% 
+  summarise(mean_unavailable = round(mean(unavailable_beds),3), 
+            mean_fundingcap = mean(funding_capacity),
+            unavailable_rate = round((mean_unavailable / mean_fundingcap)*100, 2))
 
 unavailable_beds <- unavailable_beds[order(unavailable_beds$mean_unavailable, decreasing = TRUE), ]
-
-knitr::kable(top_n(unavailable_beds,5,unavailable_beds$mean_unavailable), 
-             align ="c", col.names= c("FSA", "Average Number of Unavailable Beds"))
+#
+top_unavailability <- unavailable_beds %>% top_n(10, mean_unavailable) %>% select(fsa, mean_unavailable, unavailable_rate)
+knitr::kable(top_unavailability, 
+             align ="c", 
+             col.names= c("FSA", "Number of Unavailable Beds", 
+                          "% of Funding Capacity Unavailable"))
 ```
@@
 From this data alone, it is impossible to conclude whether these areas initially had greater rates of homelessness, thereby leading the government to spend more funds building shelters here, or that it is the higher concentration and availability of shelters here that will attract more homeless population to migrate here.
+Despite our lack of understanding of the two variable's true relationship, the correlation alone is able to indicate that emergency services are able to effectively reach the homeless population when they are placed here.
+When the city council is making policies and investments into emergency resources, it will be imperative to balance out their prioritization of these major areas of high demand versus areas which presently have less access, but may have comparatively lower demand.
 
-Next, analysis of the average bed unavailabilities across FSAs
+Next, analysis of the average bed unavailability across different FSAs revealed two specific regions of concern, M2N and M5G, which report notably higher numbers of beds being out of service each night [@tbl-unavailabilities-by-fsa].
+This is an alarming observation worthy of further investigation, as it indicates some sort of ongoing issue with the shelters in these areas that fails to convert shelter funding to actual accessible resources for the vulnerable population.
+Unless this pattern is due to unavoidable circumstances or planned projects, such as construction for shelter expansion, the administrative policy and organizational structures here may need to be evaluated to locate potential causes.
 
 ## Implications {#sec-discussion-implications}
-The observations made from this dataset reveal that the city council's effort have seen limited efficacy.
+The observations made from this dataset reveal that the city council's effort have seen limited efficacy so far.
 Occupancy rates at emergency shelters remains alarmingly high all across Toronto with no sign of decreasing.
-Though these short-term solutions
+This indicates that the city council still needs to continue efforts to find effective long-term solutions that can solve the homelessness crisis from the root.
+Research has shown that interpersonal and community resources, in addition to subsidized housing and greater income, were associated with successful transition out of homelessness [@citeAubry].
+
+As well, consistent maintenance of these emergency services is also crucial in the meantime to support the current wellbeing of the homeless population. 
+In order to improve these emergency programs which are already stretching government budgets to the maximum, problem areas such as M2N and M5G must be investigated and tackled.
 
 ## Limitations and Next Steps {#sec-discussion-limitations}
+- not a view of the complete year bc its only september
 

### 2024-09-27 paper/paper.qmd
added graph sketches + edits, added entry id back to cleaned data

@@
 ---
-title: "Emergency Homeless Shelter Availabilities Across Toronto"
+title: "The Emergency Homeless Shelter Availability Across Toronto are Insufficient For Overwhelming Demand"
 author: "Cher Ning-Li"
@@
 
-\newpage
-# Introduction
+# Introduction {#sec-introduction}
 
 Much like many other large cities around the world, Toronto is facing a homelessness crisis that seems to be ever-increasing.
-Whether it is because of environmental factors such as rising housing prices and lack of employment opportunities, or other issues such as drug abuse and mental health struggles, the homeless population of Toronto has risen to an alarming 10,811 in May 2023 [@cbcArticle].
-The city council has recognized the urgency of this problem, declaring homelessness an emergency, and pledged their full efforts towards helping the homeless population move into safe homes [@cbcArticle].
-Since then, the city of Toronto has focussed on securing more financial support from the provincial and federal government, requesting up to $25 million to fund various housing projects.
-Notably, the Canada-Ontario Housing Benefit (COHB) aimed to move 1,600-2,000 households access and transition into permanent housing in 2023-2024 [@cbcArticle].
+Whether it is because of societal factors such as rising housing prices and lack of employment opportunities, or personal issues such as substance abuse and mental health struggles, the homeless population of Toronto has risen to an alarming 10,811 in May 2023 [@cbcArticle].
+In response, the city council has declared homelessness an emergency and pledged their full efforts towards helping the homeless population move into safe homes [@cbcArticle].
+Since then, the city has focussed on securing more financial support from the provincial and federal government, requesting up to $25 million to fund various housing projects [@cbcArticle].
 
-In addition to these long-term plans that aim to move homeless individuals into stable homes, temporary shelters are also crucial during this transition process to provide short-term safety and resources.
-Research has shown that without access to shelter services, individuals can be pushed into substance-use relapses and hospitalizations, food insecurity, institutional circuitry through criminal justice systems, and other factors of instability [@citeKerman].
-Shelters also provide useful data to track and understand the current state of the city's battle against this crisis.
+In addition to these long-term projects, temporary shelters are also crucial during this transition process to provide short-term safety and resources.
+Research has shown that without access to shelter services, individuals can be pushed into substance-use relapses and hospitalizations, food insecurity, and institutional circuitry through criminal justice systems [@citeKerman].
+These shelters also provide useful data to track and understand the current progress of the city's battle against this crisis.
 
@@
 Average occupancy rate of emergency shelters per night remains at 99%, with no significant decrease across the months [@shelterDataset].
-Despite the evident high demand, there is even a decrease, albeit small, in shelters' funding capacity over the months [@shelterDataset].
+As well, the funding capacity of shelters was observed to decrease instead of increase, despite the high demand [@shelterDataset].
 
-In this paper, emergency shelter occupancy and availability will be compared across different months as well as different regions of Toronto to search for patterns or fluctutations in demand.
+In this paper, emergency shelter occupancy and availability will be compared across different months as well as different regions of Toronto to search for patterns.
 The region of each shelter will be determined by its forward sorting area (FSA) code, or the first three digits of its postal code.
@@
 The following @sec-discussion will then go more into depth to apply context to the patterns present in the data, aiming to understand the scope and urgency of the issues at hand.
-
-\newpage
 
@@
 The data is updated daily by the Toronto Shelter & Support Services and has the Open Data License (@sec-license).
-However, one limitation it faces is that it is unaudited, therefore there are limited verification methods to ensure that the shelter programs' records are accurate to their actual situation.
+However, one limitation it faces is that it is unaudited, therefore there are limited verification methods to ensure that the shelter programs' records accurately reflect their state.
 
 All shelters in this dataset are classified as either Emergency or Transitional programs, with Transitional locations providing more specialized programming and being exclusively offered to eligible individuals [@shelterDataset].
-With consideration for the greater accessibility of Emergency programs to the entire homeless population as well as the greater number of them shelters compared to Transitional, this analysis will focus on Emergency programs only [@shelterDataset].
-As well, all recorded shelters either measure capacity based on number of beds or rooms; similarly, this analysis will focus only on bed-based capacity shelters due to its greater prevalence [@shelterDataset].
+With consideration for the greater accessibility of Emergency programs to the entire homeless population as well as the greater number of them compared to Transitional type shelters, this analysis will focus on Emergency programs only [@shelterDataset].
+As well, all recorded shelters either measure capacity based on number of beds or rooms, and this analysis will focus only on bed-based capacity shelters due to its greater prevalence [@shelterDataset].
 This means that all entries pertaining to non-Toronto based, Transitional type, or room-based capacity shelters were not considered in this analysis.
@@
 # checks occupancy and capacity over months
-occupancy_means <- data %>% select(month, occupied_beds, funding_capacity, actual_capacity) %>% group_by(month) %>% 
-  summarise(mean_occupied = mean(occupied_beds), mean_capacity = mean(funding_capacity), mean_actual = mean(actual_capacity))
+occupancy_means <- data %>% select(month, occupied_beds, funding_capacity, 
+                                   actual_capacity) %>% group_by(month) %>% 
+  summarise(mean_occupied = mean(occupied_beds), 
+            mean_capacity = mean(funding_capacity), 
+            mean_actual = mean(actual_capacity))
 
@@
   scale_colour_manual("", 
-                      values = c("Occupied Beds"="darkred", "Funding Capacity"="steelblue", 
-                                 "Actual Capacity"="darkblue")) +
+                      values = c("Occupied Beds"="darkred", 
+                                 "Funding Capacity"="darkgreen", 
+                                 "Actual Capacity"="steelblue")) +
   xlab(" ") +
@@
 \newpage
-The occupancy rate can then be broken down into the mean funding capacity, actual capacity, and real number of occupied beds per month.
+Shelter occupancy can also be viewed from the angle of mean funding capacity, actual capacity, and real number of occupied beds per month.
 The high occupancy rate observed in @tbl-occupancyrate-vs-month makes it unsurprising to see the number of occupied beds consistently match so closely to the shelters' actual capacity in @fig-occupancy-vs-month.
@@
 Another dimension to consider is the geographical region of the shelters.
-Still focussing on the mean number of occupied beds, @fig-regionoccupancy shows a rather wide spread across different FSA regions, ranging from a minimum of 18.60 to a maximum of 87.74.
-Some of the regions reporting on the higher end include M5H, M6P, and M2H, which are neighbourhoods in Old Toronto, North York, and around High Park [@citeFSA].
+Still focussing on the mean number of occupied beds, @fig-regionoccupancy shows a rather wide spread of values across different FSA regions, ranging from a minimum of 18.60 to a maximum of 87.74.
+Some of the regions reporting on the higher end include M5H, M6P, and M2H, which are neighbourhoods in Old Toronto, around High Park, and North York respectively [@citeFSA].
 ```{r}
@@
 ```{r}
+#| label: tbl-fsa-bedsoccupied
+#| tbl-cap: Top 5 FSAs by Average Beds Occupied/Night
 #| echo: false
-#| include: false
-#| eval: true
 
-ordered_region_occupancy <- region_occupancy[order(region_occupancy$mean_occupied, decreasing = TRUE), ]
+ordered_region_occupancy <- 
+  region_occupancy[order(region_occupancy$mean_occupied, decreasing = TRUE), ] %>% 
+  top_n(5, mean_occupied)
+
+knitr::kable(ordered_region_occupancy, 
+             align ="c", 
+             col.names= c("FSA", "Mean Number of Beds Occupied"))
 ```
@@
 #| eval: true
-#| tbl-cap: "Number of Emergency Shelter Locations Reported in Each Toronto FSA"
+#| tbl-cap: "Top 6 FSAs by Number of Emergency Shelter Locations"
 #| echo: false
@@
 top_fsa_shelters <- fsa_shelters %>% top_n(3, numshelters)
-knitr::kable(top_fsa_shelters, align ="c", col.names= c("FSA", "Number of Shelter Locations"))
+
+knitr::kable(top_fsa_shelters, 
+             align ="c", 
+             col.names= c("FSA", "Number of Shelter Locations"))
 ```
@@
 
-unavailable_beds <- data %>% select(fsa, unavailable_beds, funding_capacity) %>% group_by(fsa) %>% 
+unavailable_beds <- data %>% 
+  select(fsa, unavailable_beds, funding_capacity) %>% 
+  group_by(fsa) %>% 
   summarise(mean_unavailable = round(mean(unavailable_beds),3), 
@@
 
-unavailable_beds <- unavailable_beds[order(unavailable_beds$mean_unavailable, decreasing = TRUE), ]
-#
-top_unavailability <- unavailable_beds %>% top_n(10, mean_unavailable) %>% select(fsa, mean_unavailable, unavailable_rate)
+unavailable_beds <- 
+  unavailable_beds[order(unavailable_beds$mean_unavailable, decreasing = TRUE), ]
+
+top_unavailability <- unavailable_beds %>% top_n(10, mean_unavailable) %>% 
+  select(fsa, mean_unavailable, unavailable_rate)
+
 knitr::kable(top_unavailability, 
@@
 
-As well, consistent maintenance of these emergency services is also crucial in the meantime to support the current wellbeing of the homeless population. 
+In the meantime, consistent maintenance of these emergency services is also crucial to support the current wellbeing of the homeless population.
+Temporary shelters, though unable to solve the core of the problem, are vital to preventing substance abuse relapses, suicidality, and further reduction of a support network and community connections [@citeKerman].
 In order to improve these emergency programs which are already stretching government budgets to the maximum, problem areas such as M2N and M5G must be investigated and tackled.
@@
 ## Limitations and Next Steps {#sec-discussion-limitations}
-- not a view of the complete year bc its only september
+One major limitation of the current dataset is that it only covers data from January to September 2024 as of the writing of this paper.
+This means that any patterns present in the winter months of November and December are not observable to the current analysis.
+As well, due to the dataset being unaudited, its accuracy to the true situations at these shelter locations cannot be verified.
+
+Further research should consider focussing in on the areas of higher occupancy and higher unavailabilities to understand the factors that may be causing these patterns, allowing for policy makers to better address the exact needs of the homeless population.
 
@@
 # Appendix {#sec-appendix}
+
+## Dataset and Graph Sketches {#sec-sketches}
+Sketches created when planning out the desired dataset and graphs for the current paper are included in the Github repository.
+
 ## Data License Information {#sec-license}
 "Contains information licensed under the Open Government Licence – Toronto" [@tphlicense].
-# Additional data details
+
+## Data Cleaning {#sec-datacleaning}
+@tbl-sampledata displays 5 rows of the cleaned data used for analysis in this paper.
+The process involved filtering for only columns with variables of interest, removing rows with negative or NA entries, and renaming the columns for clarity and concision.
+Additional columns for month and FSA were also created for ease of grouping during analysis.
+Packages `tidyverse` [@citeTidyverse] and `janitor` [@citeJanitor] were used.
+
+```{r}
+#| label: tbl-sampledata
+#| eval: true
+#| tbl-cap: "First 5 Rows of Cleaned Data"
+#| echo: false
+
+sample_data <- data %>% top_n(-5, id) %>% 
+  select(month, date, fsa, location_id, actual_capacity, funding_capacity, 
+         occupied_beds, unoccupied_beds, unavailable_beds)
+knitr::kable(sample_data, 
+             align = "c",
+             col.names = c("Month", "Date", "FSA", "Location ID", 
+                           "Actual Capacity", "Funding Capacity", 
+                           "Occupied Beds", "Unoccupied Beds", 
+                           "Unavailable Beds"))
+```
 

### 2024-09-27 scripts/02-data_cleaning.R

@@ -32,7 +32,7 @@ filtered$fsa <- substr(filtered$LOCATION_POSTAL_CODE, start=1, stop=3)
 
 
 # removes unnecessarily variables
-filtered <- filtered %>% select(month, OCCUPANCY_DATE, fsa, LOCATION_ID, 
+filtered <- filtered %>% select('_id', month, OCCUPANCY_DATE, fsa, LOCATION_ID, 
                                 CAPACITY_ACTUAL_BED, CAPACITY_FUNDING_BED, 
                                 OCCUPIED_BEDS, UNOCCUPIED_BEDS, 
                                 UNAVAILABLE_BEDS, OCCUPANCY_RATE_BEDS)

### 2024-09-27 paper/paper.qmd
final minor edits, grammar and citation

@@
 Dividing analysis of the data into different neighbourhoods, areas in Old Toronto are identified to have higher bed occupancies and number of shelters.
-Similarly, problem areas in North York and Old Toronto with patterns of inability to provide the number of bed spaces promised are also identified.
-Combining with the context of the current literature researching homelessness, these findings indicate a high urgency for the Toronto city council to implement effective short-term and long-term support programs to target the current homelessness crisis."
+As well, problem areas in North York and Old Toronto with patterns of inability to provide the number of bed spaces promised are identified.
+Combining with the context of the current literature on homelessness, these findings indicate a high urgency for the Toronto city council to implement effective short-term and long-term support programs to target the current homelessness crisis."
 format: pdf
@@
 Sorting for unique shelter ID's across the 32 FSA regions in this data set, @tbl-fsa-numshelters and @fig-numshelters-vs-fsa show that regions most commonly have only 1-2 shelters each, but there are areas with particularly high density.
-Specifically, the region M5A located in Old Toronto [@citeFSA] has 7 shelters, which is several times the number placed in other regions.
+Specifically, the region M5A located in Old Toronto [@citeFSA] has 7 shelters, which is several times the number available in other regions.
 
@@
 As shown previously in @fig-occupancy-vs-month, there is often a gap between a shelter's funding capacity and its actual capacity.
-This difference, reported as the number of unavailable beds in this dataset, represents resources that should have been available to the homeless population but was not on a particular night, either due to maintenance, renovations, outbreaks, or pest control issues [@shelterDataset].
+This difference, reported as the number of unavailable beds in this dataset, represents resources that should have been available to the homeless population but was not, either due to maintenance, renovations, outbreaks, or pest control issues [@shelterDataset].
 
 Using similar methods as above to group data by FSA code, the average number of unavailable bed spaces per region is found to be 1.30.
-
 Despite this promisingly low number, @tbl-unavailabilities-by-fsa displays some of the FSAs with highest average unavailability per night and shows that the distribution across regions is noticeably skewed.
@@
 ```{r}
-#| eval: true
+#| eval: false
 #| include: false
@@
 
-summary(unavailable_beds)
+# summary(unavailable_beds)
 ```
@@
 Analysis of this dataset in @sec-data-findings yielded several conclusions about the still-prevalent homelessness crisis here in Toronto.
-First, the consistently high occupancy rate throughout the year [@tbl-occupancyrate-vs-month] suggests that the need for housing resources remains high, and is not particularly affected by time of year.
-However, as observed in @fig-occupancy-vs-month, there is a relatively steep dip in funding capacity around March that is not preceded by a noticeable dip in demand.
+First, the consistently high occupancy rate [@tbl-occupancyrate-vs-month] suggests that the need for housing resources remains high and is not particularly affected by time of year.
+However, as observed in @fig-occupancy-vs-month, there is a relatively steep dip in funding capacity around March that is not preceded by any noticeable dip in demand.
 This decrease is potentially due to the city emphasizing their efforts during winter months, as those are intuitively the most dangerous months for those living on the streets [@cbcArticle].
@@
 
-Moving onto FSA-based observations, it was found that there are several neighbourhoods including around Old Toronto and North York that have some of the highest concentration of emergency shelters and have highest funding capacities [@tbl-fsa-numshelters, @fig-numshelters-vs-fsa].
-Despite the high occupancy [@fig-regionoccupancy] in these specific areas, it is important to remember that these findings are purely correlational and not causational.
+Moving onto FSA-based observations, it was found that there are several neighbourhoods, including some around Old Toronto and North York, that report the highest concentrations of emergency shelters and the highest funding capacities (@tbl-fsa-numshelters).
+Despite the high occupancy in these specific areas, as shown in [@fig-regionoccupancy], it is important to remember that these findings are purely correlational and not causational.
 From this data alone, it is impossible to conclude whether these areas initially had greater rates of homelessness, thereby leading the government to spend more funds building shelters here, or that it is the higher concentration and availability of shelters here that will attract more homeless population to migrate here.
@@
 
-Next, analysis of the average bed unavailability across different FSAs revealed two specific regions of concern, M2N and M5G, which report notably higher numbers of beds being out of service each night [@tbl-unavailabilities-by-fsa].
+Next, analysis of the average bed unavailability across different FSAs revealed two specific regions of concern, M2N and M5G, which report notably higher numbers of beds being out of service each night (@tbl-unavailabilities-by-fsa).
 This is an alarming observation worthy of further investigation, as it indicates some sort of ongoing issue with the shelters in these areas that fails to convert shelter funding to actual accessible resources for the vulnerable population.
@@
 The observations made from this dataset reveal that the city council's effort have seen limited efficacy so far.
-Occupancy rates at emergency shelters remains alarmingly high all across Toronto with no sign of decreasing.
+Occupancy rates at emergency shelters remain alarmingly high all across Toronto with no sign of decreasing.
 This indicates that the city council still needs to continue efforts to find effective long-term solutions that can solve the homelessness crisis from the root.
 Research has shown that interpersonal and community resources, in addition to subsidized housing and greater income, were associated with successful transition out of homelessness [@citeAubry].
+In addition to continuing to gather funds for affordable housing projects, this may represent another direction worthy of consideration for the city.
 
@@
 This means that any patterns present in the winter months of November and December are not observable to the current analysis.
-As well, due to the dataset being unaudited, its accuracy to the true situations at these shelter locations cannot be verified.
+As well, due to the dataset being unaudited, its accuracy to the true situations at these shelter locations cannot be fully verified.
 
@@
 
-sample_data <- data %>% top_n(-5, id) %>% 
-  select(month, date, fsa, location_id, actual_capacity, funding_capacity, 
-         occupied_beds, unoccupied_beds, unavailable_beds)
+sample_data <- data %>% top_n(-5, id)
 knitr::kable(sample_data, 
              align = "c",
-             col.names = c("Month", "Date", "FSA", "Location ID", 
+             col.names = c("id", "Month", "Date", "FSA", "Location ID", 
                            "Actual Capacity", "Funding Capacity", 
                            "Occupied Beds", "Unoccupied Beds", 
-                           "Unavailable Beds"))
+                           "Unavailable Beds", "Occupancy Rate"))
 ```

### 2024-09-27 paper/paper.qmd
final edits

@@
 ---
-title: "The Emergency Homeless Shelter Availability Across Toronto are Insufficient For Overwhelming Demand"
+title: "The Emergency Homeless Shelter Availability Across Toronto is Insufficient For Overwhelming Demand"
 author: "Cher Ning-Li"
@@
 
-With that in mind, how did the city council's efforts turn out?
+With that in mind, how has the city council's efforts turned out thus far?
 Analyzing Toronto's shelter occupancy data from 2024, we do not see optimistic results.
@@
 All shelters in this dataset are classified as either Emergency or Transitional programs, with Transitional locations providing more specialized programming and being exclusively offered to eligible individuals [@shelterDataset].
-With consideration for the greater accessibility of Emergency programs to the entire homeless population as well as the greater number of them compared to Transitional type shelters, this analysis will focus on Emergency programs only [@shelterDataset].
-As well, all recorded shelters either measure capacity based on number of beds or rooms, and this analysis will focus only on bed-based capacity shelters due to its greater prevalence [@shelterDataset].
+With consideration for the greater accessibility of Emergency programs to the entire homeless population as well as the greater number of them compared to Transitional type shelters, this analysis will focus on Emergency programs only.
+As well, all recorded shelters either measure capacity based on number of beds or rooms, and this analysis will focus only on bed-based capacity shelters due to its greater prevalence.
 This means that all entries pertaining to non-Toronto based, Transitional type, or room-based capacity shelters were not considered in this analysis.
-It should also be noted that as this dataset contains only data from this year, only data from January 1 to September 26 was accessible as of writing this paper, and therefore only these entries were utilized in analysis.
+It should also be noted that as this dataset covers only 2024, only the entries between January 1 and September 26 were accessible as of writing this paper, and therefore only these data points were utilized in analysis.
 
@@
 Funding capacity refers to the maximum number of beds the location is able to offer, and represents the sum of actual capacity and unavailable beds.
-Unavailable beds refers to the number of spaces that are out of service due to logistical reasons [@shelterDataset].
-Actual capacity represents the number of beds that are usable per night, and should be the sum of occupied and unoccupied beds.
+Unavailable beds refers to the number of spaces that are out of service due to logistical reasons .
+Actual capacity represents the number of beds that are usable per night, and should be the sum of occupied and unoccupied beds [@shelterDataset].
 After the dataset was cleaned for entries with empty values and invalid negative values, all entries in the remaining set were shown to adhere to these relations when tested.
@@
 Packages `tidyverse` [@citeTidyverse] and `janitor` [@citeJanitor] were then used to clean and test the downloaded dataset.
-The R code used to clean, test, and display data were adapted from @citeTB.
+Much of the R code used to clean, test, and display data were adapted from @citeTB.
 
@@
 \newpage
-Shelter occupancy can also be viewed from the angle of mean funding capacity, actual capacity, and real number of occupied beds per month.
+Shelter occupancy can also be viewed from the angle of mean funding capacity, actual capacity, and number of occupied beds per month.
 The high occupancy rate observed in @tbl-occupancyrate-vs-month makes it unsurprising to see the number of occupied beds consistently match so closely to the shelters' actual capacity in @fig-occupancy-vs-month.
@@
 Another dimension to consider is the geographical region of the shelters.
-Still focussing on the mean number of occupied beds, @fig-regionoccupancy shows a rather wide spread of values across different FSA regions, ranging from a minimum of 18.60 to a maximum of 87.74.
+Still focussing on the mean number of occupied beds, @fig-regionoccupancy shows a rather wide spread of values across different FSA regions, ranging from a minimum of 19 to a maximum of 88.
 Some of the regions reporting on the higher end include M5H, M6P, and M2H, which are neighbourhoods in Old Toronto, around High Park, and North York respectively [@citeFSA].
@@
 #| label: tbl-fsa-bedsoccupied
-#| tbl-cap: Top 5 FSAs by Average Beds Occupied/Night
+#| tbl-cap: Top 5 FSAs by Average Beds Occupied Per Night
 #| echo: false
 
+region_occupancy$mean_occupied <- round(region_occupancy$mean_occupied, 0)
 ordered_region_occupancy <- 
@@
 
-Using similar methods as above to group data by FSA code, the average number of unavailable bed spaces per region is found to be 1.30.
+Using similar methods as above to group data by FSA code, the average number of unavailable bed spaces per region is found to be 1.
 Despite this promisingly low number, @tbl-unavailabilities-by-fsa displays some of the FSAs with highest average unavailability per night and shows that the distribution across regions is noticeably skewed.
 M2N and M5G especially stand out, representing neighbourhoods in North York and Old Toronto respectively [@citeFSA].
+
 The unavailability as a percentage of funding capacity was also calculated to verify whether this gap could be simply proportional to the total capacity in these regions, and results indicate that this is not the case.
@@
 The implications of this finding will be discussed further in @sec-discussion.
+All numbers in @tbl-unavailabilities-by-fsa are rounded for the appropriate number of significant figures.
 
@@
   group_by(fsa) %>% 
-  summarise(mean_unavailable = round(mean(unavailable_beds),3), 
+  summarise(mean_unavailable = round(mean(unavailable_beds),0), 
             mean_fundingcap = mean(funding_capacity),
-            unavailable_rate = round((mean_unavailable / mean_fundingcap)*100, 2))
+            unavailable_rate = round((mean_unavailable / mean_fundingcap)*100, 0))
 
@@
 Analysis of this dataset in @sec-data-findings yielded several conclusions about the still-prevalent homelessness crisis here in Toronto.
-First, the consistently high occupancy rate [@tbl-occupancyrate-vs-month] suggests that the need for housing resources remains high and is not particularly affected by time of year.
+First, the consistently high occupancy rate (@tbl-occupancyrate-vs-month) suggests that the need for housing resources remains high and is not particularly affected by time of year.
 However, as observed in @fig-occupancy-vs-month, there is a relatively steep dip in funding capacity around March that is not preceded by any noticeable dip in demand.

### 2024-09-27 scripts/00-simulate_data.R

@@ -31,10 +31,13 @@ simulated_data <-
       ),
       origin = "1970-01-01"
     ),
-    capacity = rpois(n = number_of_dates, lambda = 50),
-    occupied = min(capacity, rpois(n = number_of_dates, lambda = 50)),
-    unoccupied = capacity - occupied,
-    occupancy_rate = (occupied/capacity) *100
+    actual_capacity = rpois(n = number_of_dates, lambda = 50),
+    funding_capacity = max(rpois(n = number_of_dates, lambda = 50), 
+                           rpois(n = number_of_dates, lambda = 60)),
+    occupied = min(actual_capacity, rpois(n = number_of_dates, lambda = 50)),
+    unoccupied = actual_capacity - occupied,
+    unavailable = funding_capacity - actual_capacity,
+    occupancy_rate = round((occupied/actual_capacity) *100, 2)
   )
 
 # Write csv

### 2024-09-27 paper/paper.qmd
minor writing edits

@@
 
-Moving onto FSA-based observations, it was found that there are several neighbourhoods, including some around Old Toronto and North York, that report the highest concentrations of emergency shelters and the highest funding capacities (@tbl-fsa-numshelters).
-Despite the high occupancy in these specific areas, as shown in [@fig-regionoccupancy], it is important to remember that these findings are purely correlational and not causational.
+Moving onto FSA-based observations, it was found that there are several neighbourhoods, including some around Old Toronto and North York, that report the highest numbers of emergency shelter locations and the highest funding capacities (@tbl-fsa-numshelters).
+At first glance, the high occupancy also observed in the areas with higher funding seems to suggest that these regions experience more severe homelessness.
+However, it is important to remember that these findings are purely correlational and not causational.
 From this data alone, it is impossible to conclude whether these areas initially had greater rates of homelessness, thereby leading the government to spend more funds building shelters here, or that it is the higher concentration and availability of shelters here that will attract more homeless population to migrate here.
-Despite our lack of understanding of the two variable's true relationship, the correlation alone is able to indicate that emergency services are able to effectively reach the homeless population when they are placed here.
+Despite our lack of understanding of the two variables' true relationship, the correlation alone is able to indicate that emergency services are able to effectively reach the homeless population when they are placed here.
 When the city council is making policies and investments into emergency resources, it will be imperative to balance out their prioritization of these major areas of high demand versus areas which presently have less access, but may have comparatively lower demand.
@@
 ## Limitations and Next Steps {#sec-discussion-limitations}
-One major limitation of the current dataset is that it only covers data from January to September 2024 as of the writing of this paper.
+One major limitation of the current dataset is that it only covers data from January to September 26, 2024 as of the writing of this paper.
 This means that any patterns present in the winter months of November and December are not observable to the current analysis.
@@
 
-sample_data <- data %>% top_n(-5, id)
+sample_data <- data %>% top_n(-5, id) %>% 
+  select(month, date, fsa, location_id, actual_capacity, funding_capacity, 
+         occupied_beds, unoccupied_beds, unavailable_beds, occupancy_rate)
 knitr::kable(sample_data, 
              align = "c",
-             col.names = c("id", "Month", "Date", "FSA", "Location ID", 
+             col.names = c("Month", "Date", "FSA", "Location ID", 
                            "Actual Capacity", "Funding Capacity", 
